---
title: "Figure 02 Short Chain Fatty Acids"
author: "Paul J. McMurdie"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: rmdformats::readthedown
params:
  tabMetabolonScfaLong: "../tidy_data/tabMetabolonScfaLong.RDS"
  tabScfaEventWide: "../tidy_data/tabMetabolonScfaEventWide.RDS"
  tabSfu: "../tidy_data/tabSfu.RDS"
  tabFecalScfa: "../tidy_data/tabFecalScfa.RDS"
  tabPp: "../tidy_data/tabPp.RDS"
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_chunk$set(
  warning = FALSE,
  error = TRUE, 
  echo = TRUE, 
  fig.width = 10, 
  fig.height = 7)
```

```{r}
library("magrittr")
library("data.table")
library("ggplot2"); packageVersion("ggplot2")
library("patchwork"); packageVersion("patchwork")
library("GGally"); packageVersion("GGally")
library("ggcorrplot"); packageVersion("ggcorrplot")
library("ggbeeswarm"); packageVersion("ggbeeswarm")
```

```{r}
theme_set(theme_bw())
scaleColorManualTreatment <-
  scale_color_manual(
    values = 
      c(
        placebo = "gray50",
        wbf10 = "darkblue",
        wbf11 = "darkgreen"
      )
  )
scaleFillManualTreatment <-
  scale_fill_manual(
    values = 
      c(
        placebo = "gray50",
        wbf10 = "darkblue",
        wbf11 = "darkgreen"
      )
  )
studyArmPretty <- 
  c(placebo = "Placebo", 
    wbf10 = "WBF-010", 
    wbf11 = "WBF-011")
scaleColorTreatmentManPretty <-
  scale_color_manual(
    values = 
      c(
        "Placebo" = "gray50",
        "WBF-010" = "darkblue",
        "WBF-011" = "darkgreen"
      )
  )
scaleFillTreatmentManPretty <-
  scale_fill_manual(
    values = 
      c(
        "Placebo" = "gray50",
        "WBF-010" = "darkblue",
        "WBF-011" = "darkgreen"
      )
  )
scfaNamePretty <-
  c("Acetic acid" = "Acetate", 
    "Propionic acid" = "Propionate", 
    "Butyric acid" = "Butyrate")
vecOrderScfa <- 
  c("Acetic acid", "Propionic acid", "Butyric acid") %>% 
  rev()
```


# Load data

```{r}
tabMetabolonScfaLong <- readRDS(params$tabMetabolonScfaLong)
tabScfaEventWide <- readRDS(params$tabScfaEventWide)
tabSfu <- readRDS(params$tabSfu)
tabFecalScfa <- readRDS(params$tabFecalScfa)
# Per protocol
tabPp <- readRDS(params$tabPp)
```

Show participants per arm represented in main specimen collection.

```{r}
tabMetabolonScfaLong %>% 
  .[, .(No.Participants = uniqueN(Subject)), by = "treatment"] %>% 
  knitr::kable(digits = 3)
```



# Fig 1A: Delta SCFA

## Prepare table

```{r}
tabKey <- NULL
tabKey <-
  tabMetabolonScfaLong %>% copy %>% 
  .[c("Acetic acid", "Propionic acid", "Butyric acid"), on = "BIOCHEMICAL"] %>% 
  # Set observations from the same subject+time to the mean among replicates
  # There are not many of these...
  .[, ConcMicroMolePerLiter := mean(ConcMicroMolePerLiter), 
    by = .(BIOCHEMICAL, Subject, Event)] %>% 
  .[, .SD[1], by = .(BIOCHEMICAL, Subject, Event)]
tabKey %>% nrow()
tabKey %>% .$BIOCHEMICAL %>% unique()
# Define subject factor order based on butyrate concentration at baseline
orderSubjectBaselineButyrate <-
  tabKey %>% copy %>% 
  .[(BIOCHEMICAL == "Butyric acid")] %>% 
  .[(Event == "Baseline")] %>% 
  setorder(treatment, ConcMicroMolePerLiter) %>% 
  .$Subject
tabKey[, subjectFactor := factor(x = Subject, levels = orderSubjectBaselineButyrate)]
```

## SCFA at baseline

```{r}
tabKeyScfaBaseline <-
  tabKey %>% copy %>% 
  .[(Event == "Baseline")] %>% 
  .[, Treatment := studyArmPretty[treatment]]
# Compute the medians
tabKeyScfaBaseline %>% 
      .[, .(
        GrandMedian = median(ConcMicroMolePerLiter, na.rm = TRUE),
        stdDev = sd(ConcMicroMolePerLiter, na.rm = TRUE)
      ), by = "BIOCHEMICAL"] %>% 
  knitr::kable(digits = 1)

tabKeyScfaBaseline %>% 
  ggplot(aes(Treatment, ConcMicroMolePerLiter, color = Treatment)) +
  facet_wrap(~BIOCHEMICAL, scales = "free_y") +
  geom_hline(
    linetype = 3,
    data = tabKeyScfaBaseline %>% 
      .[, .(GrandMedian = median(ConcMicroMolePerLiter, na.rm = TRUE)), 
        by = "BIOCHEMICAL"],
    mapping = aes(yintercept = GrandMedian), 
    show.legend = FALSE) +
  geom_boxplot(
    fill = NA,
    outlier.shape = NA
  ) +
  geom_point(
    size = 3,
    stroke = 0,
    position = position_beeswarm(cex = 4, groupOnX = TRUE)
  ) +
  scaleColorTreatmentManPretty +
  theme(
    panel.grid = element_blank(),
    axis.ticks.x = element_blank(),
    strip.background = element_blank(),
    legend.position = "none", 
    )
```


## $\Delta$ Butyrate Concentration

```{r}
tabScfaDeltas <- tabWinGrpDelta <- tabBwGrpDelta <- NULL
tabScfaDeltas <-
  tabKey %>% 
  # unfortunate case-complexity
  .[, Treatment := copy(treatment)] %>% 
  # Compute concentration deltas
  dcast.data.table(
    formula = Treatment + Subject + BIOCHEMICAL ~ Event, 
    value.var = "ConcMicroMolePerLiter", 
    fun.aggregate = mean, 
    na.rm = TRUE) %>% 
  .[, deltaConcentration := Week12 - Baseline]

# Within-group, research question: Did it increase?
tabWinGrpDeltaGreater <- 
  tabScfaDeltas %>% 
  .[, .(
    wilcoxOut = list(
      wilcox.test(
        x = deltaConcentration, 
        alternative = "greater",
        conf.int = TRUE,
        mu = 0.0, 
        paired = FALSE)
    )
  ), 
  by = .(Treatment, BIOCHEMICAL)]

# Collect the full confidence interval for plotting
tabWinGrpDelta <- 
  tabScfaDeltas %>% 
  .[, .(
    wilcoxOut = list(
      wilcox.test(
        x = deltaConcentration, 
        alternative = "two.sided",
        conf.int = TRUE,
        mu = 0.0, 
        paired = FALSE)
    )
  ), 
  by = .(Treatment, BIOCHEMICAL)]
# Extract key results from test
extract_wi = function(wilcoxOut){data.table(
  pvalue = wilcoxOut$p.value,
  statistic = wilcoxOut$statistic,
  estimate = wilcoxOut$estimate,
  confIntLo = wilcoxOut$conf.int[1],
  confIntHi = wilcoxOut$conf.int[2]
)}
tabWinGrpDelta <- tabWinGrpDelta[, extract_wi(wilcoxOut[[1]]), 
                                 by = .(Treatment, BIOCHEMICAL)]
tabWinGrpDeltaGreater <- 
  tabWinGrpDeltaGreater[, extract_wi(wilcoxOut[[1]]), 
                        by = .(Treatment, BIOCHEMICAL)]
setorder(tabWinGrpDelta, Treatment, -pvalue)
setorder(tabWinGrpDeltaGreater, Treatment, -pvalue)
# show table
tabWinGrpDelta %>% knitr::kable(digits = 3)
tabWinGrpDeltaGreater %>% knitr::kable(digits = 3)
# What percent increase is this for butyrate?
medianButBaseline <-
  tabKey[(BIOCHEMICAL == "Butyric acid" & 
            Treatment == "wbf11" & 
            Event == "Baseline"), 
         median(ConcMicroMolePerLiter)]
# This is the percent increase from the median baseline concentration:
(
  (medianButBaseline + 
     tabWinGrpDelta[(BIOCHEMICAL == "Butyric acid" & Treatment == "wbf11")]$estimate) / 
    medianButBaseline - 1) %>% 
  multiply_by(100) %>% 
  round(digits = 0)
  
# Between-group test on deltas
tabBwGrpDelta <-
  tabScfaDeltas %>% 
  .[(Treatment %in% c("wbf11", "placebo"))] %>% 
  .[, .(
    wilcoxOut = list(
      try(expr = {
        wilcox.test(
          x = deltaConcentration[(Treatment == "wbf11")],
          y = deltaConcentration[(Treatment == "placebo")],
          alternative = "greater",
          conf.int = TRUE,
          paired = FALSE)
      }, silent = TRUE)
    )
  ), 
  by = .(BIOCHEMICAL)]
# Extract key results from test
extract_wi = function(wilcoxOut){
  data.table(
    pvalue = wilcoxOut$p.value,
    statistic = wilcoxOut$statistic,
    estimate = wilcoxOut$estimate,
    confIntLo = wilcoxOut$conf.int[1],
    confIntHi = wilcoxOut$conf.int[2]
  )
}
# Check that test didn't fail and return non-sense (missing or funky data)
tabBwGrpDelta[, success := inherits(wilcoxOut[[1]], "htest"), 
              by = .(BIOCHEMICAL)]
tabBwGrpDelta <-
  tabBwGrpDelta %>% 
  .[(success), try({extract_wi(wilcoxOut[[1]])}, silent = TRUE), 
    by = .(BIOCHEMICAL)]
setorder(tabBwGrpDelta, -pvalue)
tabBwGrpDelta %>% knitr::kable(digits = 3)

# Summarize SCFA changes, clear butyrate increase.
padAnnotation = 0.2
tabFig2a <-
  tabWinGrpDelta %>% copy %>% 
  .[, Treatment := studyArmPretty[Treatment]] %>% 
  .[, scfa := factor(scfaNamePretty[BIOCHEMICAL], levels = rev(scfaNamePretty))]
# molecule-specific limits to enforce centering across panels despite free scale.
tabFig2aLims <- 
  tabFig2a %>% copy %>% 
  .[, .(
    y = (1 + padAnnotation) * 
      max(abs(confIntHi), abs(confIntLo))
  ), by = .(Treatment, scfa)]
pFig2aConcentrationDelta <-
  tabFig2a %>% 
  ggplot(aes(Treatment, estimate)) +
  facet_wrap(~scfa, scales = "free_y") +
  geom_hline(yintercept = 0, size = 0.1) +
  geom_blank(
    data = tabFig2aLims,
    mapping = aes(
      y = y,
      ymin = -y,
      ymax = y)
  ) +
  geom_pointrange(
    # alpha = 1.0,
    # size = 1.25,
    size = 1.,
    fatten = 4,
    shape = 23,
    stroke = 0,
    # fill = "black",
    mapping = aes(
      fill = Treatment,
      color = Treatment,
      # alpha = confIntLo > 0,
      # fill = confIntLo > 0,
      # color = confIntLo > 0,
      y = estimate,
      ymin = confIntLo,
      ymax = confIntHi)
  ) +
  scaleColorTreatmentManPretty +
  scaleFillTreatmentManPretty +
  # Bracket annotations
  ggpubr::geom_bracket(
    size = 0.2,
    bracket.nudge.y = -0.03,
    vjust = 0.7,
    hjust = 1,
    tip.length = c(0.3, 0.002),
    data = data.table(
      scfa = factor(
        x = "Butyrate", 
        levels = scfaNamePretty %>% rev())
    ),
    xmin = "Placebo", 
    xmax = "WBF-011", 
    y.position = 0.37, 
    label = "*") +
  geom_text(
    data = data.table(
      scfa = factor(
        x = "Butyrate", 
        levels = scfaNamePretty %>% rev())
    ),
    x = "WBF-011", 
    y = 0.305,
    label = "**"
  ) +
  ylab("\u0394 Concentration [\u03BCM]") +
  # ggtitle("\u0394 Concentration [\u03BCM]") +
  theme(
    plot.title = element_blank(),
    strip.background.x = element_rect(fill = NA, color = NA),
    strip.text.x = element_text(
      hjust = 0,
      size = 10,
      # margin = margin(0, 0, 1.4, 0, "pt"),
      face = "plain"
    ),
    panel.border = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(size = 0.2),
    legend.position = "none",
    axis.text = element_text(size = 9),
    axis.ticks.x = element_blank(),
    axis.ticks.y = element_blank(),
    # axis.ticks.y = element_line(size = 0.2),
    axis.title.y = element_text(size = 12),
    axis.title.x = element_blank()
  )
pFig2aConcentrationDelta
```


# Fig 1B: Correlation bw fecal and circulating SCFA

## Join faecal and plasma SCFA

Show anti-intersections.
Note: There are some ITT-only participants included in stool data.
Loss of a few plasma-pairs (due to labeling, etc)
on the plasma side.

```{r}
setdiff(tabFecalScfa$Subject, tabKey$Subject) %>% length()
setdiff(tabKey$Subject, tabFecalScfa$Subject) %>% length()
```

Join

```{r}
tabFecalAndPlasma <- NULL
tabFecalAndPlasma <-
  (tabFecalScfa %>% copy %>% 
     setnames("Concentration", "fecal_mM")) %>% 
  .[(tabKey %>% copy %>% setnames("ConcMicroMolePerLiter", "plasma_uM")), 
    on = c("Subject", "Event", "BIOCHEMICAL")]
```

Single trendline for each split.
Let axes be free because acetate 
is canonically much higher concentration.

```{r}
tabFecalAndPlasma %>% copy %>% 
  # join to add SFU status
  (tabSfu[, .(SulfonylureaDetected = SulfonylureaDetected[1]), 
          by = "Subject"])[., on = "Subject"] %>% 
  ggplot(aes(fecal_mM, plasma_uM)) +
  # facet_wrap(BIOCHEMICAL ~ Event, scales = "free") +
  facet_grid(BIOCHEMICAL ~ paste(Treatment, Event, sep = ": "), 
             scales = "free") +
  # Enforce zero is in y-axis range
  geom_hline(
    size = 0.2,
    yintercept = 0.0,
    color = "black",
    alpha = 0.75) +
  geom_point(mapping = aes(color = SulfonylureaDetected), show.legend = FALSE) +
  scale_color_manual(values = c("TRUE" = "gray", "FALSE" = "black")) +
  geom_smooth(
    size = 0.25, 
    # mapping = aes(color = treatment),
    se = FALSE,
    color = "black",
    method = robustbase::lmrob)
```

Is there a positive association there for butyrate?

Assess correlation, lm

```{r}
tabFecalPlasmaLmRes <- NULL
tabFecalPlasmaLmRes <- 
  tabFecalAndPlasma %>% 
  .[!is.na(fecal_mM)] %>% 
  .[, .(
    lmrobOut = list(
      robustbase::lmrob(
        formula = plasma_uM ~ fecal_mM,
        data = .SD,
        na.action = na.exclude)
    ),
    spearmanCor = cor(fecal_mM, plasma_uM, method = "spearman"),
    spearmanCorPval = 
      cor.test(fecal_mM, plasma_uM, method = "spearman")$p.value
  ), by = .(BIOCHEMICAL, Treatment, Event)]


extract_lmrob = function(lmrobOut){
  x <- NULL
  x <- lmrobOut %>% summary()
  data.table(
    slope = x$coefficients["fecal_mM", "Estimate"],
    slopeP = x$coefficients["fecal_mM", "Pr(>|t|)"],
    slopeSE = x$coefficients["fecal_mM", "Std. Error"]
  )
}

# (robust) lm results, butyrate at week-12 stands out
# As it does by eyeball in the chart
tabFecalPlasmaLmResVals <- NULL
tabFecalPlasmaLmResVals <-
  tabFecalPlasmaLmRes %>% 
  .[, extract_lmrob(lmrobOut[[1]]), 
    by = .(BIOCHEMICAL, Treatment, Event)] %>% 
  setorder(BIOCHEMICAL, Treatment, Event)
tabFecalPlasmaLmResVals

# Butyrate stands out here
tabFecalPlasmaLmResVals %>% 
  .[(slope > 0 & slopeP < 0.1)] %>% 
  knitr::kable(digits = 3)

# And correlation? Not as significant in assessment.
tabFecalPlasmaLmRes %>% copy %>% 
  .[, lmrobOut := NULL] %>% 
  setorder(BIOCHEMICAL, Treatment, Event) %>% 
  knitr::kable(digits = 3)
```

Plot polish

```{r}
pFecalPlasmaFit <- pFecalPlasmaFitSimple <- tabFecalAndPlasmaPlot <- NULL
tabFecalAndPlasmaPlot <-
  tabFecalAndPlasma %>% copy %>% 
  .[(BIOCHEMICAL == "Butyric acid")] %>% 
  .[, Treatment := factor(x = studyArmPretty[treatment], 
                          levels = studyArmPretty)] %>%
  # Create new panel label from Treatment and Event
  .[, armTimeChar := paste(Treatment, Event, sep = ": ")] %>%  
  setorder(Treatment, Event) %>% 
  .[, armTime := factor(armTimeChar, levels = unique(armTimeChar))] %>% 
  # For plotting purposes, set plasma values above 1.5 to Inf
  # Will have to explain these were left outside range as possible outliers
  # or at least skewing the rest of the pattern observable
  .[(plasma_uM > 1.5), plasma_uM := Inf]
pFecalPlasmaFitSimple <-
  tabFecalAndPlasmaPlot %>% 
  ggplot(aes(fecal_mM, plasma_uM)) +
  # Omit extra bit of the scale
  scale_y_continuous(expand = c(0, 0)) +
  # facet_grid(treatment ~ Event, scales = "free") +
  facet_wrap(facets = vars(armTime), nrow = 1) +
  # scaleColorManualTreatment +
  # Enforce zero is in y-axis range
  geom_hline(
    size = 0.5,
    yintercept = 0.0,
    color = "black",
    alpha = 0.75) +
  # Dummy fit to emphasize the lack of pattern elsewhere
  geom_smooth(
    data = tabFecalAndPlasmaPlot,
    color = "darkgray",
    se = FALSE,
    size = 0.35,
    fullrange = TRUE,
    na.rm = TRUE,
    method = robustbase::lmrob,
  ) +
  # Just the non-trivial fit
  geom_smooth(
    data = tabFecalAndPlasmaPlot %>% 
      .[(Treatment == "WBF-011")] %>% 
      .[(Event == "Week12")],
    color = "darkgreen",
    fill = "gray90",
    size = 0.35,
    fullrange = TRUE,
    na.rm = TRUE,
    method = robustbase::lmrob,
  ) +
  geom_point(
    na.rm = TRUE,
    size = 0.4,
    color = "black",
    # mapping = aes(color = treatment),
    show.legend = FALSE) +
  ylab("Plasma butyrate [\u03BCM]") +
  xlab("Fecal butyrate [mM]") +
  theme(
    strip.background = element_rect(fill = NA, color = NA),
    strip.text.x = element_text(
      hjust = 0.5,
      size = 8,
      margin = margin(0, 0, 1.4, 1, "pt")
    ),
    panel.border = element_rect(size = 0.2),
    panel.spacing.y = unit(5, units = "pt"),
    panel.spacing.x = unit(5, units = "pt"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    legend.position = "none",
    axis.text = element_text(size = 9),
    axis.ticks = element_line(size = 0.2),
    axis.title = element_text(size = 11)
  )
# pFecalPlasmaFitSimple

# For main plot, add p-value annotation.
pFecalPlasmaFit <-
  pFecalPlasmaFitSimple +
  # Annotate slope and p-value
  geom_text(
    x = 9,
    y = 0.75,
    size = 2,
    mapping = aes(label = paste0("p = ", round(slopeP, digits = 3))),
    data =
      tabFecalPlasmaLmResVals %>% copy %>%
      .[, Treatment := factor(x = studyArmPretty[Treatment],
                              levels = studyArmPretty)] %>%
      .[, armTimeChar := paste(Treatment, Event, sep = ": ")] %>%
      .[, armTime := factor(armTimeChar, levels = unique(armTimeChar))] %>% 
      .[(slope > 0 & slopeP < 0.1)]
  )
  # ggtitle("Association b/w fecal and circulating butyrate at Week-12?")
pFecalPlasmaFit
```


# Fig 1C: $\Delta$ HbA1c v. $\Delta$ Butyrate

Correlation changes in HbA1c and plasma SCFA

## Join 

```{r}
tabGluCtlScfa <-
  tabPp %>% 
  dcast.data.table(Subject ~ Event, value.var = c("A1C", "AUC_TOT_GLU")) %>% 
  .[, deltaA1c := A1C_Week12 - A1C_Baseline] %>% 
  .[, .(Subject, deltaA1c)] %>% 
  # Join with scfa table
  .[tabScfaEventWide, on = "Subject"]
```

## Prepare param tables for Fig 1C

Show fit if SFU-detected are considered.
$\Delta$A1c v. $\Delta$Butyrate

```{r}
# SFU+ subjects excluded
tabGluCtlScfa %>% 
  .[(treatment == "wbf11")] %>% 
  .[!(SulfonylureaDetected)] %>%
  .[(BIOCHEMICAL == "Butyric acid")] %>% 
  robustbase::lmrob(formula = deltaA1c ~ Week12MinusBaseline,
                    data = .) %>% 
  summary()
# SFU+ subjects excluded, 
# high-leverage (high butyrate, < -2 deltaA1c) excluded
# (for the sake of argument)
tabGluCtlScfa %>% 
  .[(treatment == "wbf11")] %>% 
  .[!(SulfonylureaDetected)] %>%
  .[(BIOCHEMICAL == "Butyric acid")] %>% 
  .[(deltaA1c > -2)] %>% # 0.008
  robustbase::lmrob(formula = deltaA1c ~ Week12MinusBaseline,
                    data = .) %>% 
  summary()

# SFU param included rather than SFU+ excluded
tabGluCtlScfa %>% 
  .[(treatment == "wbf11")] %>% 
  .[(BIOCHEMICAL == "Butyric acid")] %>% 
  robustbase::lmrob(formula = deltaA1c ~ SulfonylureaDetected + Week12MinusBaseline,
                    data = .) %>% 
  summary()
```

The $\Delta$A1c v. $\Delta$Butyrate relationship is stronger 
than the SFU-naive and log-ratio fit would make it appear.
Log-ratio is mostly useful in this context when multiple SCFA shown on the same plot
(especially acetate dominates in both concentration baseline and change).
Since we have already shown how butyrate stands out 
(especially in for negative-slope, when SFU is considered),
and are showing just butyrate in the main figure,
it makes sense to use the *natural* unit $\Delta$Butyrate [uM].

Correlation has a weak p-value (but WBF-011 most significant ranking).

```{r}
tabGluCtlScfa %>% 
  # .[!(SulfonylureaDetected)] %>%
  .[(BIOCHEMICAL == "Butyric acid")] %>% 
  .[!is.na(deltaA1c)] %>% 
  .[, .(
    spearmanCor = 
      cor(Week12MinusBaseline, deltaA1c, method = "spearman"),
    spearmanCorPval = 
      cor.test(Week12MinusBaseline, deltaA1c, method = "spearman")$p.value
  ), by = .(BIOCHEMICAL, treatment, SulfonylureaDetected)] %>% 
  setorder(treatment, SulfonylureaDetected) %>%
  knitr::kable(digits = 2)
```

Define function to run lm function,
extract results into table
or return `NULL` if something went wrong

```{r}
run_lm_fun = function(testFormula, dt, lmfun = robustbase::lmrob){
  linmod <- dtOut <- linModSummary <- NULL
  trash <-
    try(silent = TRUE, expr = {
      linmod <- 
        lmfun(
          formula = testFormula,
          data = dt)
    })
  if(!is.null(linmod)){
    trash <- try(silent = TRUE, expr = {
      dtOut <-
        summary(linmod) %>% 
        coefficients() %>% 
        as.data.frame() %>%
        set_colnames(c("paramEstimate", "stdErr", "tValue", "pFitParam")) %>% 
        data.table(keep.rownames = "CoefficientName", skip_absent=TRUE) %>% 
        .[, r2 := (summary(linmod)$r.squared)]
    })
  }
  return(dtOut)
}
```

The linear model slope parameter very negative,
stands out dramatically from the others.

```{r}
tabGluCtlDeltaButyrate <- NULL
tabGluCtlDeltaButyrate <-
  tabGluCtlScfa %>% 
  .[(BIOCHEMICAL == "Butyric acid")] %>% 
  .[, .(
    lmA1c = list(
      run_lm_fun(
        testFormula = 
          formula(deltaA1c ~ Week12MinusBaseline,
                  env = tabGluCtlScfa),
        dt = .SD, 
        lmfun = robustbase::lmrob)
    )
  ), by = .(BIOCHEMICAL, treatment, SulfonylureaDetected)]
# Unfurl params
tabGluCtlDeltaButyrateParams <-
  tabGluCtlDeltaButyrate %>% 
  .[!(lmA1c %>% sapply(is.null))] %>% 
  .[, data.table(lmA1c[[1]]),
    by = .(BIOCHEMICAL, treatment, SulfonylureaDetected)] %>% 
  setorder(-r2, pFitParam)
tabGluCtlDeltaButyrateParams %>% 
  .[, .(BIOCHEMICAL, treatment, SulfonylureaDetected,
        CoefficientName, paramEstimate, pFitParam, r2)] %>% 
  knitr::kable(digits = 3)
# Since it is drops below the digit threshold in the table,
# Show it explicitly 
# (also shown above in the standard model summary):
tabGluCtlDeltaButyrateParams[1]$pFitParam
```

Define plot.

```{r}
tabDeltaGluCtlButPlot <- pDeltaGluCtlBut <- NULL
tabDeltaGluCtlButPlot <-
  tabGluCtlScfa %>% copy %>% 
  .[!is.na(deltaA1c)] %>% 
  .[(BIOCHEMICAL == "Butyric acid")] %>% 
  .[, Treatment := factor(x = studyArmPretty[treatment], 
                          levels = studyArmPretty)]

# with WBF-011 high-leverage point excluded
tabDeltaGluCtlButHighLev <-
  tabDeltaGluCtlButPlot %>% copy %>% 
  .[(deltaA1c < -2)]
tabDeltaGluCtlButPlot <-
  tabDeltaGluCtlButPlot %>% 
  .[(deltaA1c > -2)]
  
# Define plot
pDeltaGluCtlBut <-
  tabDeltaGluCtlButPlot %>% 
  ggplot(aes(Week12MinusBaseline, deltaA1c)) +
  facet_wrap(~Treatment, nrow = 1) +
  geom_hline(yintercept = 0.0, size = 0.1) +
  geom_vline(xintercept = 0.0, size = 0.1) +
  geom_smooth(
    data = tabDeltaGluCtlButPlot %>% 
      .[(Treatment == "WBF-011" & !SulfonylureaDetected)],
    mapping = aes(group = SulfonylureaDetected),
    method = robustbase::lmrob, 
    color = "darkgreen",
    fill = "gray90",
    size = 0.35,
    fullrange = TRUE,
    na.rm = TRUE,
    se = TRUE) +
  geom_text(
    hjust = 0,
    x = 0.75, 
    y = -0.5,
    size = 2.5,
    # Format
    # parse = FALSE,
    # label = "p < 10^-6",
    # .[(deltaA1c > -2)] %>% # 0.008
    # See model summaries above, 
    # this is with high-leverage point omitted.
    label = "p = 0.008",
    # mapping = aes(label = pFitParam),
    data = tabGluCtlDeltaButyrateParams %>% copy %>% 
      .[(CoefficientName == "Week12MinusBaseline")] %>% 
      .[(treatment == "wbf11")] %>%
      .[(SulfonylureaDetected == FALSE)] %>% 
      .[, Treatment := factor(x = studyArmPretty[treatment], 
                              levels = studyArmPretty)]
  ) +
  geom_point(
    mapping = aes(fill = SulfonylureaDetected),
    shape = 21,
    size = 2, 
    stroke = 0
  ) + 
  # Add back the high-leverage participant as open-circle
  geom_point(
    data = tabDeltaGluCtlButHighLev, 
    shape = 21,
    size = 1.5, 
    fill = "white",
    stroke = 0.5
  ) +
  scale_fill_manual(values = c("TRUE" = "gray", "FALSE" = "black")) +
  xlab("\u0394 Plasma butyrate [\u03BCM]") +
  ylab("\u0394 HbA1c") +
  theme(
    axis.title.y = element_text(size = 12, angle = 0, vjust = 0.5),
    axis.title.x = element_text(size = 10),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 10),
    strip.background.x = element_rect(fill = NA, color = NA),
    strip.background.y = element_rect(fill = NA),
    legend.position = "none",
    axis.ticks = element_blank()
    )
# pDeltaGluCtlBut
```

## Fig 1c: $\Delta$A1c v. $\Delta$Butyrate

Add in-figure legend

```{r}
pDeltaGluCtlBut <-
  pDeltaGluCtlBut +
  # Label highest and lowest points as in-chart legend
  ggrepel::geom_text_repel(
    seed = 711,
    segment.size = 0.2,
    box.padding = 0.5,
    point.padding = 0.1,
    nudge_x = 1,
    nudge_y = 0.2,
    data = pDeltaGluCtlBut$data %>% copy %>% 
      .[, thisLegend := NA_character_] %>% 
      # only write the labels on WBF-011 (right)
      .[(treatment == "wbf10")] %>% 
      .[(deltaA1c == max(deltaA1c[which(SulfonylureaDetected == TRUE)], 
                         na.rm = TRUE)), 
        thisLegend := "SFU use"] %>% 
      .[(deltaA1c == min(deltaA1c[which(SulfonylureaDetected == FALSE)], 
                         na.rm = TRUE)), 
        thisLegend := "No SFU use"] %>% 
      .[!is.na(thisLegend)],
    mapping = aes(label = thisLegend, 
                  color = SulfonylureaDetected),
    min.segment.length = 0,
    size = 3
  ) +
  scale_color_manual(values = c("TRUE" = "gray", "FALSE" = "black"))
pDeltaGluCtlBut
```



# Fig 1D: delta heatmap

## Prepare $\Delta$ tables

```{r}
tabPpDelta <- NULL
tabPpDelta <-
  tabPp %>% copy %>% 
  .[(Event %in% c("Baseline", "Week12"))] %>% 
  melt.data.table(
    id.vars = c("treatment", "Subject", "Event"), 
    variable.name = "clinicMeasure", 
    value.name = "clinicValue") %>% 
  dcast.data.table(
    formula = treatment + Subject + clinicMeasure ~ Event, 
    value.var = "clinicValue") %>% 
  .[, deltaValue := Week12 - Baseline] %>% 
  # prepend varnames so that "delta" is in front
  .[, ClinMeasDelta := paste0("delta", clinicMeasure)] %>% 
  dcast.data.table(
    formula = Subject ~ ClinMeasDelta,
    value.var = "deltaValue")
#   # Compute the deltas
tabPpDelta %>% show()
```

## $\Delta$ Corr Heatmap

```{r}
# Define inclusion and order
gluChar <- c(
  "deltaAUC_INC_INS",
  "deltaAUC_TOT_INS", 
  "deltaFASTING_INSULIN",
  "deltaFASTING_GLUCOSE",
  "deltaAUC_TOT_GLU",
  "deltaAUC_INC_GLU",
  "deltaA1C"
) %>% 
  rev()

scfaChar <- 
  c(
    "Acetic acid", 
    "Propionic acid", 
    "Butyric acid") 
deltaScfaNamePretty <-
  c("Acetic acid" = "deltaAcetate", 
    "Propionic acid" = "deltaPropionate", 
    "Butyric acid" = "deltaButyrate")
showChar <- c(scfaChar, gluChar)
orderChar <- c(deltaScfaNamePretty[scfaChar] %>% unname(), gluChar)
# Convert all 'delta' in variable names to the unicode greek delta symbol
# "\u0394"
orderChar <- 
  gsub("^delta", "\u0394 ", orderChar) %>% 
  set_names(orderChar)
# Standardize the nomenclature for glycated hemoglobin
orderChar <- gsub("A1C", "HbA1c", orderChar)
# Standardize capitalization (not all lower case)
orderChar <- gsub("FASTING", "Fasting", orderChar)
orderChar <- gsub("_INSULIN", " insulin", orderChar)
orderChar <- gsub("_INS$", " insulin", orderChar)
orderChar <- gsub("_GLUCOSE", " glucose", orderChar)
orderChar <- gsub("_GLU$", " glucose", orderChar)
orderChar <- gsub("_TOT", "_tot", orderChar)
orderChar <- gsub("_INC", "_inc", orderChar)

# Define change correlation table
tabDeltaCorr <- NULL
tabDeltaCorr <-
  tabScfaEventWide %>%
  .[(BIOCHEMICAL %in% scfaChar)] %>% 
  # cast wide on Subject
  dcast.data.table(
    formula = treatment + Subject ~ BIOCHEMICAL, 
    value.var = "log2Wk12OverBaseline", 
    drop = TRUE, 
    fun.aggregate = mean, 
    na.rm = TRUE) %>% 
  # Join to add study values
  .[(tabPpDelta), on = "Subject", nomatch = 0] %>% 
  # Remove the categoricals, except for treatment
  .[, -c("Subject"), with = FALSE] %>% 
  .[, .(
    spearmat = list(spearman = cor(x = .SD, method = "spearman", 
                                   use = "pairwise.complete.obs")),
    spearmat.p = list(cor_pmat(x = .SD, method = "spearman",
                               use = "pairwise.complete.obs"))
  ), by = .(treatment)]

tabDeltaCorrVals <-
  tabDeltaCorr %>% 
  .[,
    (
      data.table(spearmat[[1]], keep.rownames = "row") %>% 
        melt.data.table(id.vars = "row", 
                        variable.name = "col", 
                        value.name = "Correlation")
    ), by = "treatment"] %>% 
  # Join with p-value matrix
  .[(
    tabDeltaCorr %>% 
      .[, (data.table(spearmat.p[[1]], keep.rownames = "row") %>% 
         melt.data.table(id.vars = "row", 
                         variable.name = "col", 
                         value.name = "p")), 
        by = "treatment"]),
    on = c("treatment", "row", "col")]

# Define plotting table
tabSimpleDeltaCorr <-  NULL
tabSimpleDeltaCorr <- 
  tabDeltaCorrVals %>% copy %>% 
  # Filter
  .[((row %in% showChar) & (col %in% showChar))] %>% 
  # Pretty arm names
  .[, Treatment := factor(x = studyArmPretty[treatment], 
                          levels = studyArmPretty)]
# Unify the SCFA names
tabSimpleDeltaCorr[, Row := row %>% as.character() %>% copy()]
tabSimpleDeltaCorr[(row %in% names(deltaScfaNamePretty)), 
                   Row := deltaScfaNamePretty[row]]
tabSimpleDeltaCorr[, Col := col %>% as.character() %>% copy()]
tabSimpleDeltaCorr[(Col %in% names(deltaScfaNamePretty)), 
                   Col := deltaScfaNamePretty[Col]]

# Create a facet-version of row, col to enforce ordering.
tabSimpleDeltaCorr[, Row := factor(orderChar[Row], levels = orderChar)]
tabSimpleDeltaCorr[, Col := factor(orderChar[Col], levels = orderChar)]


## define $\Delta$ heatmap
pHeatDeltaSimple <- NULL
pHeatDeltaSimple <- 
  tabSimpleDeltaCorr %>% copy %>% 
  # Significance simplifier for plotting
  .[(p > 0.4), Correlation := 0.0] %>% 
  ggplot(mapping = aes(Row, Col, fill = Correlation)) +
  facet_wrap(facets = vars(Treatment), nrow = 1) +
  geom_raster() +
  scale_fill_gradient2(
    low = "darkcyan",
    mid = "white",
    high = "chocolate4",
    midpoint = 0,
    breaks = c(-1, -0.5, 0, 0.5, 1)
  ) +
  # Adjust the color mapping legend
  guides(
    fill = guide_colourbar(
      # nbin = 5, 
      barwidth = 20, 
      barheight = 0.5)
  ) +
  theme(
    legend.box = "horizontal",
    legend.direction = "horizontal",
    # legend.text = element_text(vjust = 0.7),
    legend.title = element_text(vjust = 1),
    legend.position = "bottom",
    strip.background.x = element_rect(fill = "white", color = "black"),
    strip.text.x = element_text(size = 10),
    # strip.background.y = element_rect(fill = NA),
    axis.title = element_blank(),
    # plot.background = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    panel.grid = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    # axis.text.x = element_text(angle = 75, hjust = 1),
    axis.text.y = element_text(size = 10, hjust = 0, margin = margin(r = -3))
  )
## Show heatmap
pHeatDeltaSimple
```


# Build Figure 02 

Combined figure summarizing these results...

```{r pFig2, fig.width=8.5, fig.height=9}
pFig2 <- NULL
fig2Layout <- "
A
B
C
D
"
pFig2 <-
  (
    pFig2aConcentrationDelta + 
      theme(axis.title.y = element_text(margin = margin(r = -110, unit = "pt")))
  ) +
  # pFecalPlasmaFit /
  (
    pFecalPlasmaFit + 
      theme(axis.title.y = element_text(margin = margin(r = -110, unit = "pt")))
  ) +
  (
    pDeltaGluCtlBut + 
      theme(axis.title.y = element_text(margin = margin(r = -80, unit = "pt")))
  ) +
  # 1012
  pHeatDeltaSimple +
  plot_layout(design = fig2Layout) +
  plot_annotation(tag_levels = 'a') & 
  theme(plot.tag = element_text(size = 10, face = "bold"))
pFig2
```

```{r ggsave}
ggsave("Figure-02.png", pFig2, width = 8.5, height = 9)
```



