---
title: "Figure 04 Untargeted Metabolomics"
author: "Paul J. McMurdie"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: rmdformats::readthedown
params:
  tabMetabolonUntgtLong: "../tidy_data/tabMetabolonUntgtLong.RDS"
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_chunk$set(
  warning = FALSE,
  error = TRUE, 
  echo = TRUE, 
  fig.width = 10, 
  fig.height = 7)
```

```{r}
library("magrittr")
library("data.table")
library("ggplot2"); packageVersion("ggplot2")
library("ggridges"); packageVersion("ggridges")
library("FactoMineR"); packageVersion("FactoMineR")
library("factoextra"); packageVersion("factoextra")
library("patchwork"); packageVersion("patchwork")
library("ggbeeswarm"); packageVersion("ggbeeswarm")
# library("ggrepel"); packageVersion("ggrepel")
```

```{r}
theme_set(
  theme_bw() +
    theme(
      panel.grid = element_blank(), 
      axis.ticks = element_line(size = 0.25),
      strip.background = element_blank(),
      strip.text.y = element_text(angle = 0)
    )
)
scaleColorManualTreatment <-
  scale_color_manual(
    values = 
      c(
        placebo = "gray50",
        wbf10 = "darkblue",
        wbf11 = "darkgreen"
      )
  )
scaleFillManualTreatment <-
  scale_fill_manual(
    values = 
      c(
        placebo = "gray50",
        wbf10 = "darkblue",
        wbf11 = "darkgreen"
      )
  )
studyArmPretty <- 
  c(placebo = "Placebo", 
    wbf10 = "WBF-010", 
    wbf11 = "WBF-011")
scaleColorTreatmentManPretty <-
  scale_color_manual(
    values = 
      c(
        "Placebo" = "gray50",
        "WBF-010" = "darkblue",
        "WBF-011" = "darkgreen"
      )
  )
scaleFillTreatmentManPretty <-
  scale_fill_manual(
    values = 
      c(
        "Placebo" = "gray50",
        "WBF-010" = "darkblue",
        "WBF-011" = "darkgreen"
      )
  )
scfaNamePretty <-
  c("Acetic acid" = "Acetate", 
    "Propionic acid" = "Propionate", 
    "Butyric acid" = "Butyrate")
vecOrderScfa <- 
  c("Acetic acid", "Propionic acid", "Butyric acid") %>% 
  rev()
```


# Load data

```{r}
# long-form table of the untargeted metabolomics data
tabMetabolonUntgtLong <- readRDS(params$tabMetabolonUntgtLong)
tabMetabolonUntgtLong %>% dim()
tabMetabolonUntgtLong$COMP_ID %>% uniqueN()
```

## Log-ratio

Compute subject-wise Week12, Baseline paired metrics

```{r}
tabPivEventWide <-
  tabMetabolonUntgtLong %>% 
  # pivot wide to ensure consistent handling 
  # of subject-event-sample and missing
  dcast.data.table(
    formula = treatment + Subject + COMP_ID + BIOCHEMICAL ~ Event, 
    fill = NA, 
    fun.aggregate = mean,
    value.var = "ScaledImpData") %>% 
  # subtract week12 from baseline
  .[, Week12MinusBaseline := Week12 - Baseline] %>% 
  .[, log2Wk12OverBaseline := log2(Week12 / Baseline)]
```

```{r}
tabPivEventWideOrigScale <-
  tabMetabolonUntgtLong %>% 
  # pivot wide to ensure consistent handling 
  # of subject-event-sample and missing
  dcast.data.table(
    formula = treatment + Subject + COMP_ID + BIOCHEMICAL ~ Event, 
    fill = NA, 
    # This is needed due to two subjects having a replicate timepoint
    fun.aggregate = mean,
    value.var = "OrigScale") %>% 
  .[, log2RatioOrigScale := log2(Week12 / Baseline)]
```

Join to add to `tabPivEventWide`.
This adds explicit `NA` that was 'missing' in `ScaledImpData`.

```{r}
tabPivEventWide <-
  tabPivEventWide %>% 
  .[tabPivEventWideOrigScale[, .(Subject, COMP_ID, BIOCHEMICAL, log2RatioOrigScale)], 
    on = .(Subject, COMP_ID, BIOCHEMICAL)]
```

## Missing values

Evaluate, flag, entries that have high number of missing data in an arm,
or high number of artificial no-change in an arm (log-ratio is zero).

```{r}
# Tally sums of non-missing values
tabPivEventWide[, numNonMissing := sum(!is.na(log2RatioOrigScale)), 
                     by = .(treatment, BIOCHEMICAL)]
# Tally sums of non-zero log-ratios
tabPivEventWide[, numNonzeroLogRatio := 
                  sum(log2Wk12OverBaseline != 0.0, na.rm = TRUE),
                by = .(treatment, BIOCHEMICAL)]
```

## Check for differences between arms at baseline

```{r}
tabMetabolonUntgtLong %>% 
  .[(Event == "Baseline")] %>% 
  .[, .(medianSubject = median(ScaledImpData, na.rm = TRUE)), 
    by = .(treatment, BIOCHEMICAL)] %>% 
  # Compute differences with placebo
  .[, .(wbf11 = medianSubject[treatment == "wbf11"] -
          medianSubject[treatment == "placebo"], 
        wbf10 = medianSubject[treatment == "wbf10"] -
          medianSubject[treatment == "placebo"]), 
    by = "BIOCHEMICAL"] %>% 
  melt.data.table(id.vars = "BIOCHEMICAL", 
                  variable.name = "treatment", 
                  variable.factor = FALSE) %>% 
  .[, Treatment := studyArmPretty[treatment]] %>% 
  ggplot(aes(x = value, y = Treatment, fill = Treatment)) +
  geom_blank(mapping = aes(x = -value)) +
  # geom_vline(xintercept = 0.0, size = 0.2, color = "gray") +
  geom_density_ridges(scale = 1.6, rel_min_height = 0.0, alpha = 1.0) +
  # scale_y_discrete(expand = expansion(mult = c(0.0, 0))) +
  scaleFillTreatmentManPretty +
  theme_ridges() +
  # xlim(-2.5, 2.5) +
  xlab("Difference between medians at baseline")
```

Also see PCA of baseline-only values, below.


# Fig. 3a-c

Most metabolites aren't different at baseline and don't change

# Fig. 3a median log-ratio distributions

Distribution of medians by arm and biochemical

```{r}
tabMedianDists <-
  tabPivEventWide %>% 
  .[!is.na(log2RatioOrigScale)] %>% 
  .[!is.na(log2Wk12OverBaseline)] %>% 
  .[, .(medianLog2Wk12OverBaseline = median(log2Wk12OverBaseline, na.rm = TRUE)), 
    by = .(treatment, BIOCHEMICAL)]
```

ggridges summary of distribution of median log-ratios.

```{r}
pLogRatioRidges <-
  tabMedianDists %>% copy %>% 
  .[, Treatment := studyArmPretty[treatment]] %>% 
  ggplot(aes(x = medianLog2Wk12OverBaseline, y = Treatment, fill = Treatment)) +
  # geom_vline(xintercept = 0.0, size = 0.2, color = "gray") +
  geom_density_ridges(scale = 1.6, rel_min_height = 0.0, alpha = 1.0) +
  scale_y_discrete(expand = expansion(mult = c(0.0, 0))) +
  scaleFillTreatmentManPretty +
  xlim(-2.5, 2.5) +
  # xlab("median Log2-Ratio") +
  ggtitle(expression(paste(
    "Median ",
    log[2](over(Endpoint, Baseline))
  ))) +
  theme_ridges(font_size = 13, grid = TRUE) +
  theme(
    legend.position = "none",
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.title = element_text(hjust = 0.5, size = 8),
    # axis.title.x = element_text(hjust = 0.5),
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )
pLogRatioRidges
```


# Fig. 3b QQ Plot

Compare them by arm more directly.

```{r}
munge_quantiles = function(xx, probSeq = seq(0, 1, by = 0.005)){
  yy = quantile(
    x = xx,
    probs = probSeq,
    na.rm = TRUE)
  return(
    data.table(
      quantChar = names(yy), 
      percentile = names(yy) %>% gsub("\\%", "", .) %>% as.numeric(),
      value = yy)
  )
}
tabMedianDistQs <- NULL
tabMedianDistQs <-
  tabMedianDists %>% 
  .[, munge_quantiles(
    xx = medianLog2Wk12OverBaseline
  ), by = .(treatment)] %>% 
  dcast.data.table(quantChar + percentile ~ treatment, value.var = "value")
```

## Fig. 3b QQ Plot

Combined qqplot for build

```{r}
pQqLogRatio <-
  tabMedianDistQs %>% 
  melt.data.table(
    id.vars = c("quantChar", "percentile", "placebo"), 
    variable.name = "studyArm", 
    variable.factor = FALSE, 
    value.factor = FALSE,
    value.name = "log2Ratio") %>% 
  setorder(studyArm, percentile) %>% 
  ggplot(aes(placebo, log2Ratio, color = studyArm)) +
  geom_abline() +
  geom_path(size = 1.5) +
  geom_point(size = 2.5, stroke = 0) +
  scaleColorManualTreatment +
  ylab("Formulation") +
  xlab("Placebo") +
  ggtitle("Q-Q plot") +
  theme(
    legend.position = 'none',
    panel.grid = element_blank())
pQqLogRatio
```


# Fig. 3c PCA

Sanity-check that there were not large 
group-wise structural differences at baseline.

```{r}
tabUntgtBaselineWide <- NULL
tabUntgtBaselineWide <-
  tabMetabolonUntgtLong %>% 
  .[(Event == "Baseline")] %>% 
  dcast.data.table(
    formula = treatment + Subject ~ COMP_ID, 
    value.var = "ScaledImpData", 
    fun.aggregate = mean, na.rm = TRUE)
tabUntgtBaselineWide %>% dim()
tabUntgtBaselineWide[1:5, 1:5, with = FALSE]
# Use factominer, factoextra for viz.
# Perform PCA
pcaBaseline <- PCA(tabUntgtBaselineWide[, -c(1, 2), with = FALSE],  graph = FALSE)
```

## Scree plot, baseline

```{r}
fviz_eig(pcaBaseline) + 
  theme(
    panel.grid.major.x = element_blank(), 
    axis.ticks = element_blank())
```

Define PCA subject plot for figure

```{r}
fviz_pca_ind(
  X = pcaBaseline, 
  label = "none", 
  habillage = tabUntgtBaselineWide$treatment %>% factor(),
  addEllipses = TRUE, 
  ellipse.level = 0.95
) +
  scaleColorManualTreatment +
  scaleFillManualTreatment +
  theme(
    legend.position = "none",
    plot.title = element_blank(),
    panel.grid = element_blank()
  ) + 
  ggtitle("PCA, baseline-only") +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid = element_blank()
  )
```

PCA on subject-wise log-ratio values.

```{r}
tabLogRatioWide <-
  tabPivEventWide %>%
  # Pivot even wider
  dcast.data.table(
    formula = treatment + Subject ~ COMP_ID, 
    value.var = "log2Wk12OverBaseline")
tabLogRatioWide[1:5, 1:5, with = FALSE]
# Perform PCA
pcaLogRatio <- PCA(tabLogRatioWide[, -c(1, 2), with = FALSE],  graph = FALSE)
```

## Scree plot, log-ratio

```{r}
fviz_eig(pcaLogRatio) + 
  theme(
    panel.grid.major.x = element_blank(), 
    axis.ticks = element_blank())
```

Define PCA subject plot for figure

```{r}
pLogRatioPCA <-
  fviz_pca_ind(
    X = pcaLogRatio, 
    label = "none", 
    habillage = tabLogRatioWide$treatment %>% factor(),
    addEllipses = TRUE, 
    ellipse.level = 0.95
  ) +
  scaleColorManualTreatment +
  scaleFillManualTreatment +
  ggtitle("PCA") +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5),
    panel.grid = element_blank()
  )
pLogRatioPCA
```


# Fig. 3d: Between-Group Volcano plot

## WBF-011 - Placebo

```{r}
tabWilcoxBwGrpWbf11Placebo <- NULL
# Compute the between-group comparison WBF-011 v. Placebo
tabWilcoxBwGrpWbf11Placebo <-
  tabPivEventWide %>% 
  .[(numNonMissing > 5)] %>% 
  .[(treatment %in% c("wbf11", "placebo"))] %>% 
  .[, .(
    wilcoxOut = list(
      try(expr = {
        wilcox.test(
          x = log2Wk12OverBaseline[(treatment == "wbf11")],
          y = log2Wk12OverBaseline[(treatment == "placebo")],
          alternative = "two.sided",
          conf.int = TRUE,
          paired = FALSE)
      }, silent = TRUE)
    )
  ), 
  by = .(BIOCHEMICAL)]
# Extract key results from test
extract_wi = function(wilcoxOut){
  data.table(
    pvalue = wilcoxOut$p.value,
    statistic = wilcoxOut$statistic,
    estimate = wilcoxOut$estimate,
    confIntLo = wilcoxOut$conf.int[1],
    confIntHi = wilcoxOut$conf.int[2]
  )
}
# Check that test didn't fail and return non-sense (missing or funky data)
tabWilcoxBwGrpWbf11Placebo[, success := inherits(wilcoxOut[[1]], "htest"), by = .(BIOCHEMICAL)]
tabWilcoxBwGrpWbf11Placebo <-
  tabWilcoxBwGrpWbf11Placebo %>% 
  .[(success), try({extract_wi(wilcoxOut[[1]])}, silent = TRUE), 
    by = .(BIOCHEMICAL)]
setorder(tabWilcoxBwGrpWbf11Placebo, -pvalue)
```

## WBF-010 - Placebo

```{r}
tabWilcoxBwGrpWbf10Placebo <- NULL
# Compute the between-group comparison WBF-011 v. Placebo
tabWilcoxBwGrpWbf10Placebo <-
  tabPivEventWide %>% 
  .[(numNonMissing > 5)] %>% 
  # .[(numNonzeroLogRatio > thresholdMinNonzeroLogRatios)] %>%
  .[(treatment %in% c("wbf10", "placebo"))] %>% 
  .[, .(
    wilcoxOut = list(
      try(expr = {
        wilcox.test(
          x = log2Wk12OverBaseline[(treatment == "wbf10")],
          y = log2Wk12OverBaseline[(treatment == "placebo")],
          alternative = "two.sided",
          conf.int = TRUE,
          paired = FALSE)
      }, silent = TRUE)
    )
  ), 
  by = .(BIOCHEMICAL)]
# Extract key results from test
extract_wi = function(wilcoxOut){
  data.table(
    pvalue = wilcoxOut$p.value,
    statistic = wilcoxOut$statistic,
    estimate = wilcoxOut$estimate,
    confIntLo = wilcoxOut$conf.int[1],
    confIntHi = wilcoxOut$conf.int[2]
  )
}
# Check that test didn't fail and return non-sense (missing or funky data)
tabWilcoxBwGrpWbf10Placebo[, success := inherits(wilcoxOut[[1]], "htest"), by = .(BIOCHEMICAL)]
tabWilcoxBwGrpWbf10Placebo <-
  tabWilcoxBwGrpWbf10Placebo %>% 
  .[(success), try({extract_wi(wilcoxOut[[1]])}, silent = TRUE), 
    by = .(BIOCHEMICAL)]
setorder(tabWilcoxBwGrpWbf10Placebo, -pvalue)
```


## Plot: Summarize comparisons

Add metabolite annotations on wilcoxon table,
then curate key pathways to highlight in plot.

```{r}
# Curate highlights
tabWilcoxBwGrpWbf11PlaceboAnnot <- NULL
tabWilcoxBwGrpWbf11PlaceboAnnot <- 
  tabMetabolonUntgtLong %>% 
  .[, .SD[1], by = "BIOCHEMICAL"] %>% 
  .[, .(BIOCHEMICAL, SUPER_PATHWAY, SUB_PATHWAY)] %>% 
  .[tabWilcoxBwGrpWbf11Placebo, on = "BIOCHEMICAL", nomatch = 0] %>% 
  copy()
setnames(tabWilcoxBwGrpWbf11PlaceboAnnot, "estimate", "estimateBwGrp")
setnames(tabWilcoxBwGrpWbf11PlaceboAnnot, "pvalue", "pBwGrp")

# First curate key pathways to highlight
tabWilcoxBwGrpWbf11PlaceboAnnot[, subPathway := NA_character_]

## FAO - Fatty Acid Oxidation, Acyl Carnitines
tabWilcoxBwGrpWbf11PlaceboAnnot[grep("Acyl Carnitine", SUB_PATHWAY)]$SUB_PATHWAY %>% 
  unique() %>% sort()
# Define key representative FAO/acylcarnitine subgroups for main plot.
vecCuratedFaoSubpaths <-
  c(
    "Fatty Acid Metabolism (Acyl Carnitine, Medium Chain)",
    "Fatty Acid Metabolism (Acyl Carnitine, Dicarboxylate)",
    "Fatty Acid Metabolism (Acyl Carnitine, Polyunsaturated)",
    "Fatty Acid Metabolism (Acyl Carnitine, Hydroxy)",
    "Fatty Acid Metabolism (Acyl Carnitine, Long Chain Saturated)",
    "Fatty Acid Metabolism (Acyl Carnitine, Monounsaturated)",
    "Medium Chain Fatty Acid",
    # "Fatty Acid, Monohydroxy",
    # "Fatty Acid, Dihydroxy",
    "Fatty Acid, Dicarboxylate",
    "Long Chain Monounsaturated Fatty Acid",
    "Long Chain Polyunsaturated Fatty Acid (n3 and n6)")
tabWilcoxBwGrpWbf11PlaceboAnnot[(SUB_PATHWAY %in% vecCuratedFaoSubpaths),
                                subPathway := "Fatty Acid Oxidation"]

# Must be within lipid super-pathway
tabWilcoxBwGrpWbf11PlaceboAnnot[(
  subPathway == "Fatty Acid Oxidation" & 
    SUPER_PATHWAY != "Lipid"), 
  subPathway := NA]

# This one appears to be elevated in juvenile diabetes, 
# PMID: 2947647 DOI: 10.1002/bms.1200131004
# But possibly derived from threonine and so not relevant to FAO.
tabWilcoxBwGrpWbf11PlaceboAnnot["2R,3R-dihydroxybutyrate", 
                                on = "BIOCHEMICAL", 
                                subPathway := NA_character_]
# Also not part of fatty acid oxidation (just look at the structure)
# Derived from amino acid, or pyridine?
# "Picolinoylglycine belongs to the class of organic compounds known as n-acyl-alpha amino acids..."
# https://hmdb.ca/metabolites/HMDB0059766
tabWilcoxBwGrpWbf11PlaceboAnnot["picolinoylglycine", 
                                on = "BIOCHEMICAL", 
                                subPathway := NA_character_]

# BCAA
# This group was investigated due to prior interest,
# but the data does not support emphasizing this group
tabWilcoxBwGrpWbf11PlaceboAnnot["Leucine, Isoleucine and Valine Metabolism", 
                                subPathway := NA_character_, 
                                on = "SUB_PATHWAY"]
vecCuratedBcaaSubpaths <-
  c("Fatty Acid Metabolism (also BCAA Metabolism)",
    "Fatty Acid, Branched")
tabWilcoxBwGrpWbf11PlaceboAnnot[(SUB_PATHWAY %in% vecCuratedBcaaSubpaths),
                                subPathway := NA_character_]
# propionylcarnitine is indicative of BCAA
tabWilcoxBwGrpWbf11PlaceboAnnot[grep("propionylcarnitine", BIOCHEMICAL), 
                                subPathway := NA_character_]

# Data supports emphasizing tryptophan metabolites
tabWilcoxBwGrpWbf11PlaceboAnnot["Tryptophan Metabolism" , 
                                subPathway := "Tryptophan", 
                                on = "SUB_PATHWAY"]

# non-tryptophan aromatic amino acids.
# This group was investigated due to prior interest,
# but the data does not support emphasizing this group
nonTrpAAA <- "Phe, Tyr"
tabWilcoxBwGrpWbf11PlaceboAnnot[grep("Tyrosine", SUB_PATHWAY), 
                                subPathway := NA_character_]
tabWilcoxBwGrpWbf11PlaceboAnnot[grep("Phenylalanine", SUB_PATHWAY), 
                                subPathway := NA_character_]

# Omit '1-carboxyethyl' metabolites (octopines) from other subgroups,
# where their presence is misleading wrt the pathway
# (they are clearly their own, correlated group)
tabWilcoxBwGrpWbf11PlaceboAnnot[grep("1-carboxyethyl", BIOCHEMICAL), 
                                subPathway := NA_character_]

grepSearchStringBilirubin <- 
  "bilirubin|glucuronate|Urobilin|urobilinogen|stercobilin|heme|biliverdin"
tabWilcoxBwGrpWbf11PlaceboAnnot[grep(grepSearchStringBilirubin, BIOCHEMICAL), 
                                subPathway := "Bilirubin"]
tabWilcoxBwGrpWbf11PlaceboAnnot[grep("Bile", SUB_PATHWAY), 
                                subPathway := "Bile Acids"]

# Define plot
pVolcanoBwGrpWbf11PlaceboBase <- pVolcanoBwGrpWbf11Placebo <- NULL
pVolcanoBwGrpWbf11PlaceboBase <- 
  ggplot(
    mapping = aes(
      x = estimateBwGrp, 
      y = -log10(pBwGrp)
    )) +
  # Horizontally center
  geom_blank(
    mapping = aes(x = -estimateBwGrp), 
    data = tabWilcoxBwGrpWbf11PlaceboAnnot %>% copy() %>% 
      .[, subPathway := NULL] %>% 
      .[, SUB_PATHWAY := NULL]
  ) +
  # Volcano background cloud of all unnamed.
  geom_point(
    # Omit faceting variable so that these show up on every panel
    data = tabWilcoxBwGrpWbf11PlaceboAnnot %>% copy() %>% 
      .[, subPathway := NULL] %>% 
      .[, SUB_PATHWAY := NULL],
    fill = "gray",
    color = "gray",
    stroke = 0,
    size = 1.5, 
    alpha = 0.25) +
  geom_vline(
    xintercept = 0.0, 
    color = "gray",
    size = 0.25
  ) +
  ylab(expression(paste(-log[10](p)))) +
  xlab("WBF-011 - Placebo") +
  labs(color = "Metabolite\nGroup") +
  ggtitle("Between-group comparison, wilcoxon test", 
          "WBF-011 - placebo.") +
  theme(
    strip.text = element_text(size = 10),
    strip.background = element_blank(),
    plot.title = element_blank(),
    plot.subtitle = element_blank(),
    panel.grid = element_blank(),
    axis.title.y = element_text(size = 13, angle = 0, vjust = 0.95),
    axis.title.x = element_text(size = 13, angle = 0, vjust = 1),
    legend.position = "none"
  ) 

vecOrderSubPaths <-
  c("Fatty Acid Oxidation", "Bilirubin", "Bile Acids", "Tryptophan")
thresholdAlphaEmphasisUntgt <- 0.1

# Prepare plot definition
pVolcanoBwGrpWbf11Placebo <-
  pVolcanoBwGrpWbf11PlaceboBase +
    # Highlight!
  # color = subPathway
  geom_point(
    data = (tabWilcoxBwGrpWbf11PlaceboAnnot[!is.na(subPathway)]),
    mapping = aes(
      size = pBwGrp < thresholdAlphaEmphasisUntgt,
      alpha = pBwGrp < thresholdAlphaEmphasisUntgt
    ),
    shape = 21,
    color = "darkgreen",
    fill = "darkgreen",
    stroke = 0.1,
    alpha = 1) +
  scale_size_manual(values = c("TRUE" = 2, "FALSE" = 1)) +
  scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.5)) +
  facet_wrap(~factor(subPathway, levels = vecOrderSubPaths), 
             nrow = 2,
             drop = TRUE)
# Peek at the FAO subgroups
vecFaoShow <-
  tabWilcoxBwGrpWbf11PlaceboAnnot %>% 
  .[(subPathway == "Fatty Acid Oxidation")] %>% 
  .$SUB_PATHWAY %>% table() %>% sort() %>% 
  .[(. > 1)]
pVolcanoBwGrpWbf11PlaceboBase +
    geom_point(
    data = (tabWilcoxBwGrpWbf11PlaceboAnnot %>% 
              .[(subPathway == "Fatty Acid Oxidation")] %>% 
              .[(abs(estimateBwGrp) > 0 | pBwGrp < 1)]),
    shape = 21,
    color = "darkgreen",
    fill = "darkgreen",
    stroke = 0.1,
    size = 2, 
    alpha = 1) +
  facet_wrap(~factor(SUB_PATHWAY), drop = TRUE)

pVolcanoBwGrpWbf11Placebo
```

Compare WBF-011 and WBF-010 in MA plot

```{r}
tabBwGrpCompare <- NULL
tabBwGrpCompare <- 
  tabWilcoxBwGrpWbf10Placebo %>% copy %>% 
  setnames("estimate", "estWbf10Pbo") %>% 
  setnames("pvalue", "pWbf10Pbo") %>% 
  .[copy(tabWilcoxBwGrpWbf11PlaceboAnnot), on = "BIOCHEMICAL"] %>% 
  .[, A := 0.5 * (abs(estimateBwGrp) + abs(estWbf10Pbo))] %>% 
  .[, M := estimateBwGrp - estWbf10Pbo]

# Show labels on bile acids...
tabBwGrpCompare %>% 
  ggplot(aes(A, M)) +
  geom_point(size = 0.5, color = "gray") +
  geom_point(
    data = tabBwGrpCompare[grep("cholate", BIOCHEMICAL)],
    color = "darkorange") +
  ggrepel::geom_text_repel(
    min.segment.length = 0,
    size = 2,
    mapping = aes(label = BIOCHEMICAL),
    data = tabBwGrpCompare %>% 
      .[grep("cholate", BIOCHEMICAL)] %>% 
      .[(A > 0.4 | abs(M) > 1.0)]) +
  # .[(A > 0.65 | abs(M) > 1.0)]) +
  ggtitle("Between-group comparison with placebo, MA plot")

# Mimic the facet panels from the main figure
tabBwGrpCompare %>% copy %>% 
  .[, subPathway := NULL] %>% 
  ggplot(aes(A, M)) +
  geom_hline(yintercept = 0.0, size = 0.2) +
  geom_point(size = 0.5, color = "gray", stroke = 0) +
  geom_point(
    stroke = 0,
    size = 1,
    data = tabBwGrpCompare[!is.na(subPathway)],
    color = "darkorange") +
  facet_wrap(~subPathway) +
  ylab("WBF-011 - WBF-010") +
  ggtitle("Between-group comparison relative to placebo, MA plot") +
  theme(panel.grid = element_blank())
```

Show WBF-010 in its usual color shading simultaneously.

```{r fig.width=10, fig.height=7}
pVolcanoBwGrpBothVsPlacebo <- NULL
pVolcanoBwGrpBothVsPlacebo <-
  pVolcanoBwGrpWbf11Placebo +
  # Highlight!
  geom_point(
    data = (tabBwGrpCompare[!is.na(subPathway)]),
    mapping = aes(
      x = estWbf10Pbo,
      y = -log10(pWbf10Pbo),
      size = pWbf10Pbo < thresholdAlphaEmphasisUntgt,
      alpha = pWbf10Pbo < thresholdAlphaEmphasisUntgt
    ),
    shape = 21,
    color = "darkblue",
    fill = "darkblue",
    stroke = 0.1,
    alpha = 1) +
  facet_wrap(~factor(subPathway, levels = vecOrderSubPaths), 
             nrow = 2,
             drop = TRUE) +
  xlab("Formulation - Placebo")
pVolcanoBwGrpBothVsPlacebo
```





# Fig 3e: Select groups

## Within-group log2(Week12 / Baseline)

```{r}
tabWilcoxLog2Wk12OverBaselResult <-
  tabPivEventWide %>% 
  # .[(numNonMissing > 0)] %>% 
  .[, .(
    wilcoxOut = list(
      wilcox.test(
        x = log2Wk12OverBaseline, 
        alternative = "two.sided",
        conf.int = TRUE,
        mu = 0.0, 
        paired = FALSE)
    )
  ), 
  by = .(treatment, BIOCHEMICAL)]
# Extract key results from test
extract_wi = function(wilcoxOut){data.table(
  pvalue = wilcoxOut$p.value,
  statistic = wilcoxOut$statistic,
  estimate = wilcoxOut$estimate,
  confIntLo = wilcoxOut$conf.int[1],
  confIntHi = wilcoxOut$conf.int[2]
)}
tabWilcoxLog2Wk12OverBaselResult <-
  tabWilcoxLog2Wk12OverBaselResult[, extract_wi(wilcoxOut[[1]]), 
                                   by = .(treatment, BIOCHEMICAL)]
setnames(tabWilcoxLog2Wk12OverBaselResult, "estimate", "estimateWinGrp")
setnames(tabWilcoxLog2Wk12OverBaselResult, "pvalue", "pWinGrp")
```

## Fig 3E, within-arm changes by metabolite group

```{r fig.width=16}
# subtract within-group C.I.s for between-group 'hits'
tabSpecialWinGrp <- NULL
tabSpecialWinGrp <-
  (tabWilcoxBwGrpWbf11PlaceboAnnot %>% copy %>% 
     .[!is.na(subPathway)] %>% 
     # Same as emphasis threshold in Fig 3D
     .[(pBwGrp < thresholdAlphaEmphasisUntgt)] %>%
     .[, .(BIOCHEMICAL, subPathway, estimateBwGrp, pBwGrp)]) %>% 
  tabWilcoxLog2Wk12OverBaselResult[., on = "BIOCHEMICAL"]
# Improve the label for bilirubin degradation products
tabSpecialWinGrp[, 
                 BIOCHEMICAL := gsub(
                   pattern = "bilirubin degradation product, ", 
                   replacement = "bilirubin deriv. ", 
                   BIOCHEMICAL)]
# Define order of metabolites to help with legibility
orderSpecialMetabs <- 
  tabSpecialWinGrp %>% 
  .[(treatment == "wbf11")] %>% 
  setorder(-estimateWinGrp) %>% 
  .$BIOCHEMICAL
tabSpecialWinGrp[, metaboFactor := factor(BIOCHEMICAL, levels = orderSpecialMetabs)]
# Set CI edges to Inf if they are outside of helpful-range
tabSpecialWinGrp[(confIntLo < -1.0), confIntLo := -Inf]
tabSpecialWinGrp[(confIntHi > 1.0), confIntHi := Inf]

pSpecialWinGrp <-
  tabSpecialWinGrp %>% copy %>% 
  # Show the bile acids result in bile-acid figure, 
  # omit here to save figure space
  .[(subPathway != "Bile Acids")] %>% 
  .[, subPathwayFac := factor(
    x = subPathway, 
    levels = vecOrderSubPaths[vecOrderSubPaths != "Bile Acids"])] %>% 
  ggplot(aes(metaboFactor, estimateWinGrp, color = treatment, fill = treatment)) +
  geom_hline(yintercept = 0.0, size = 0.1) +
  facet_grid(cols = vars(subPathwayFac), scales = 'free', space = 'free') +
  # Adding this has the effect of centering each panel
  geom_blank(mapping = aes(ymin = -confIntLo, ymax = -confIntHi)) +
  # The (within-group) log2-ratio + C.I.s
  geom_pointrange(
    alpha = 1.0,
    size = 0.5,
    fatten = 4,
    shape = 23,
    stroke = 0.1,
    position = position_dodge(width = 0.6),
    mapping = aes(ymin = confIntLo, ymax = confIntHi)
  ) +
  ylab(
    expression(paste(
      log[2](over(Endpoint,Baseline))
    ))) +
  scaleColorManualTreatment +
  scaleFillManualTreatment +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    strip.background = element_blank(),
    axis.title.y = element_text(size = 7, angle = 0, vjust = 0.5),
    axis.title.x = element_blank(),
    axis.text.x = element_text(
      size = 9,
      angle = 30, 
      hjust = 1, 
      vjust = 1)
  )
pSpecialWinGrp
```


# Table: Multiple testing

```{r}
tabSummarizeMultTest <- NULL
tabSummarizeMultTest <-
  tabBwGrpCompare %>% copy %>% 
  setnames("estimateBwGrp", "estWbf11Pbo") %>% 
  setnames("pBwGrp", "pWbf11Pbo") %>% 
  setnames("subPathway", "KeyGroup") %>% 
  # Select for displaying in table
  .[, .(BIOCHEMICAL, SUPER_PATHWAY, SUB_PATHWAY, KeyGroup, 
        estWbf10Pbo, estWbf11Pbo,
        pWbf10Pbo, pWbf11Pbo)] %>% 
  setorder(pWbf11Pbo, pWbf10Pbo)

## Show independent filtering criteria (arm-blind variance)
# Compute variance for each metabolite (using the standardized, rescaled values)
tabVarUntgt <-
  tabMetabolonUntgtLong %>% 
  .[, .(Var = var(ScaledImpData, na.rm = TRUE)), by = "BIOCHEMICAL"] %>% 
  setorder(-Var)
thresholdVarianceMin <- 1

tabSummarizeMultTestIndFilt <- NULL
tabSummarizeMultTestIndFilt <- 
  tabSummarizeMultTest %>% copy %>% 
  ## Filter to stuff worth showing
  # minimum variance threshold
  .[(tabVarUntgt[(Var > thresholdVarianceMin)]$BIOCHEMICAL), on = "BIOCHEMICAL"] %>% 
  # minimum number observations for test (test failed with NA, omit)
  .[!is.na(pWbf11Pbo)] %>% 
  # Omit the unannotated metabolites. 
  # Nothing to interpret. Can show them separately later.
  .[!is.na(SUPER_PATHWAY)]

# Compute group-wise significance values
tabSummarizeMultTestIndFilt <- 
  tabSummarizeMultTestIndFilt %>% 
  ## Compute a group-wise p-value, using medians
  # SUB_PATHWAY
  .[, 
    pSubPathWilcox := wilcox.test(
      x = estWbf11Pbo, 
      alternative = "two.sided")$p.value,
    by = .(SUB_PATHWAY)] %>% 
  # KeyGroup
  .[!is.na(KeyGroup), 
    pKeyGrpWilcox := wilcox.test(
      x = estWbf11Pbo, 
      alternative = "two.sided")$p.value, 
    by = .(KeyGroup)] %>% 
  .[is.na(pKeyGrpWilcox), pKeyGrpWilcox := 1.0]

# Compute FDR adjustment for group-wise p-values
tabSubPathFdr <-
  tabSummarizeMultTestIndFilt %>% copy %>% 
  .[, .(SUB_PATHWAY, pSubPathWilcox)] %>% 
  # Omit the entries with small group size (another indpendent filter)
  .[, grpSize := .N, by = .(SUB_PATHWAY)] %>% 
  .[(grpSize > 3)] %>% 
  .[, grpSize := NULL] %>% 
  unique() %>% 
  .[, gFDRSubPath := p.adjust(p = pSubPathWilcox)] %>% 
  .[, pSubPathWilcox := NULL]
# Append group FDR
tabSummarizeMultTestIndFilt <-
  tabSummarizeMultTestIndFilt %>% 
  tabSubPathFdr[., on = "SUB_PATHWAY"]
# Move gFDRSubPath col to the right
tabSummarizeMultTestIndFilt$SubPathFDR <- tabSummarizeMultTestIndFilt$gFDRSubPath
tabSummarizeMultTestIndFilt[, gFDRSubPath := NULL]
setcolorder(tabSummarizeMultTestIndFilt, c("BIOCHEMICAL", "SUPER_PATHWAY"))
# Render summary table into report
tabSummarizeMultTestIndFilt %>% 
  setorder(pSubPathWilcox, pKeyGrpWilcox, 
           pWbf11Pbo, pWbf10Pbo) %>% 
  DT::datatable() %>% 
  DT::formatRound(
    columns = c("estWbf10Pbo", "estWbf11Pbo", 
                "pWbf10Pbo", "pWbf11Pbo",
                "pKeyGrpWilcox", "pSubPathWilcox",
                "SubPathFDR"), 
    digits = 4)
```

For completeness, show the un-annotated results

```{r}
tabSummarizeMultTest %>% 
  # Filter for unannoted results
  .[!is.na(pWbf11Pbo)] %>% 
  # Show only the unannotated metabolites. 
  .[is.na(SUPER_PATHWAY)] %>% 
  setorder(pWbf11Pbo, pWbf10Pbo) %>% 
  # knitr::kable(digits = 3)
  DT::datatable() %>% 
  DT::formatRound(
    columns = c("estWbf10Pbo", "estWbf11Pbo", "pWbf10Pbo", "pWbf11Pbo"), 
    digits=3)
```

Show the unfiltered comparison statistics for all metabolites
that had non-trivial detection 
(minimum number of subjects defined above).

```{r}
tabSummarizeMultTest %>% 
  setorder(pWbf11Pbo) %>% 
  DT::datatable() %>% 
  DT::formatRound(
    columns = c("estWbf10Pbo", "estWbf11Pbo", "pWbf10Pbo", "pWbf11Pbo"), 
    digits=4)
```




# Build Figure 4: Untargeted Metabolomics

```{r pFig4, fig.width=6.7, fig.height=8.3}
layout <- "
ABC
DDD
DDD
EEE
"
sizeAxisTitles <- 8
sizeAxisText <- 8
pFig4 <- NULL
pFig4 <-
  (pLogRatioRidges + 
     theme(
       plot.title = element_text(size = sizeAxisTitles),
       axis.text.y = element_text(size = sizeAxisTitles),
       axis.text.x = element_text(size = sizeAxisText),
       axis.ticks.x = element_line(size = 0.5)
     )
  ) +
  (pQqLogRatio +
     theme(
       plot.title = element_text(size = sizeAxisTitles),
       axis.title.x = element_text(size = sizeAxisTitles),
       axis.title.y = element_text(size = sizeAxisTitles),
       axis.text.x = element_text(size = sizeAxisText),
       axis.text.y = element_text(size = sizeAxisText),
       axis.ticks.x = element_line(size = 0.5)
     )
  ) +
  (pLogRatioPCA +
     theme(
       plot.title = element_text(size = sizeAxisTitles),
       axis.title.x = element_text(size = sizeAxisTitles),
       axis.title.y = element_text(size = sizeAxisTitles),
       axis.text.x = element_blank(),
       axis.text.y = element_blank(),
       axis.ticks = element_blank()
     )
  ) +
  (
    pVolcanoBwGrpBothVsPlacebo + 
      theme(
        strip.text = element_text(size = sizeAxisTitles),
        axis.text.x = element_text(size = sizeAxisText),
        axis.text.y = element_text(size = sizeAxisText),
        axis.title.x = element_text(size = sizeAxisTitles),
        axis.title.y = element_text(
          size = 8, 
          angle = 0, 
          vjust = 0.5,
          margin = margin(r = -35, unit = "pt"))
      )
  ) +
  (
    pSpecialWinGrp + 
      theme(
        strip.text = element_text(size = sizeAxisTitles),
        axis.text.y = element_text(size = sizeAxisText),
        axis.text.x = element_text(
          size = 6,
          angle = 40, 
          hjust = 1, 
          vjust = 1),
        axis.title.y = element_text(
          size = 6, angle = 0, vjust = 0.5,
          margin = margin(r = -30, unit = "pt"))
      )
  ) +
  plot_annotation(tag_levels = 'a') +
  plot_layout(design = layout) & 
  theme(plot.tag = element_text(size = 8, face = "bold", family = "Sans"))
# pFig4
```

```{r ggsaves}
figHeight = 210
ggsave("Figure-04.pdf", pFig4, 
       device = cairo_pdf,
       width = 170,
       dpi = 300, 
       height = figHeight, 
       units = "mm")
```



# Write tidy within-group table

```{r writetables}
tabMetabolonUntgtWilcox <- 
  (tabWilcoxBwGrpWbf11PlaceboAnnot %>% 
     .[, .(BIOCHEMICAL, subPathway, estimateBwGrp, pBwGrp)]) %>% 
  tabWilcoxLog2Wk12OverBaselResult[., on = "BIOCHEMICAL"] %>% 
  copy()
saveRDS(tabMetabolonUntgtWilcox, "../tidy_data/tabMetabolonUntgtWilcox.RDS")
```

