---
title: "Figure 03 Bile Acids"
author: "Paul J. McMurdie"
date: '`r format(Sys.Date(), "%Y-%B-%d")`'
output: rmdformats::readthedown
params:
  tabMetabolonUntgtWilcox: "../tidy_data/tabMetabolonUntgtWilcox.RDS"
  tabPlasmaMsOmicsBileAcidsLong: "../tidy_data/tabPlasmaMsOmicsBileAcidsLong.RDS"
  tabStrainCultureMsOmicsBileAcids: "../tidy_data/tabStrainCultureMsOmicsBileAcids.RDS"
  tabStrainCultureMetabolonPivWideType: "../tidy_data/tabStrainCultureMetabolonPivWideType.RDS"
  tabStrainCultureMetabolon: "../tidy_data/tabStrainCultureMetabolon.RDS"
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_chunk$set(
  warning = FALSE,
  error = TRUE, 
  echo = TRUE, 
  fig.width = 10, 
  fig.height = 7)
```

```{r}
library("magrittr")
library("data.table")
library("ggplot2"); packageVersion("ggplot2")
library("patchwork"); packageVersion("patchwork")
library("ggbeeswarm"); packageVersion("ggbeeswarm")
```

```{r}
theme_set(
  theme_bw() +
    theme(
      panel.grid = element_blank(), 
      axis.ticks = element_line(size = 0.25),
      strip.background = element_blank(),
      strip.text.y = element_text(angle = 0)
    )
)
scaleColorManualTreatment <-
  scale_color_manual(
    values = 
      c(
        placebo = "gray50",
        wbf10 = "darkblue",
        wbf11 = "darkgreen"
      )
  )
scaleFillManualTreatment <-
  scale_fill_manual(
    values = 
      c(
        placebo = "gray50",
        wbf10 = "darkblue",
        wbf11 = "darkgreen"
      )
  )
studyArmPretty <- 
  c(placebo = "Placebo", 
    wbf10 = "WBF-010", 
    wbf11 = "WBF-011")
scaleColorTreatmentManPretty <-
  scale_color_manual(
    values = 
      c(
        "Placebo" = "gray50",
        "WBF-010" = "darkblue",
        "WBF-011" = "darkgreen"
      )
  )
scaleFillTreatmentManPretty <-
  scale_fill_manual(
    values = 
      c(
        "Placebo" = "gray50",
        "WBF-010" = "darkblue",
        "WBF-011" = "darkgreen"
      )
  )
showBileAcids <- c(
  "Cholic acid" = "CA", 
  "Chenodeoxycholic acid" = "CDCA", 
  "Ursodeoxycholic acid" = "UDCA")
```


# Load data

```{r}
# Targeted plasma bile acids
tabPlasmaMsOmicsBileAcidsLong <- readRDS(params$tabPlasmaMsOmicsBileAcidsLong)
# "negative" concentration should be set to LoD to avoid logic error
tabPlasmaMsOmicsBileAcidsLong[(Concentration < 0.0), Concentration := LOD]
# Untargeted metabolites change stats (to show the plasma bile acids)
tabMetabolonUntgtWilcox <- readRDS(params$tabMetabolonUntgtWilcox)
```


# Fig. 3a Untargeted Bile Acids

Plasma untargeted bile acids summary

```{r}
# This is both the inclusion set and order it appears in the plots
showConjuFams <- c("UDCA", "CDCA", "CA", "DCA", "LCA")
# Force reset
pBileAcidUntgt <- tabBileAcidUntgt <- NULL
## Untargeted bile acid table
tabBileAcidUntgt <- 
  tabMetabolonUntgtWilcox %>% copy %>% 
  # Show the bile acids result in bile-acid figure, omit here to save figure space
  .[(subPathway == "Bile Acids")]
# Standardize nomenclature
tabBileAcidUntgt[, Molecule := copy(BIOCHEMICAL)]
tabBileAcidUntgt[, Molecule := gsub("late", "lic acid", Molecule)]
# Define conjugate families
tabBileAcidUntgt[, ConjugateFamily := "Other"]
tabBileAcidUntgt[grep("^([Tt]auro|[Gg]lyco)*[Cc]holic acid", Molecule), 
                 ConjugateFamily := "CA"]
tabBileAcidUntgt[grep("^([Z7]-)*([Kk]eto|[Tt]auro|[Gg]lyco)*[Dd]eoxycholic acid", Molecule), 
                 ConjugateFamily := "DCA"]
tabBileAcidUntgt[grep("^([Tt]auro|[Gg]lyco)*[Cc]hen(o)*deoxycholic acid", Molecule), 
                 ConjugateFamily := "CDCA"]
tabBileAcidUntgt[grep("^([Ii]so)*([Tt]auro|[Gg]lyco)*[Uu]rsodeoxycholic acid", Molecule), 
                 ConjugateFamily := "UDCA"]
tabBileAcidUntgt[grep("[lL]ithocholic acid", Molecule), 
                 ConjugateFamily := "LCA"]
tabBileAcidUntgt[grep("[Hh]yocholic acid", Molecule), 
                 ConjugateFamily := "HCA"]
tabBileAcidUntgt[grep("[Hh]yodeoxycholic acid", Molecule), 
                 ConjugateFamily := "HDCA"]
tabBileAcidUntgt[grep("muri", Molecule, ignore.case = TRUE), 
                 ConjugateFamily := "MCA"]
tabBileAcidUntgt[, OrderBA := "Z"]
tabBileAcidUntgt[(ConjugateFamily %in% c("CA", "CDCA", "HCA")), OrderBA := "Primary"]
tabBileAcidUntgt[(ConjugateFamily %in% c("DCA", "LCA", "UDCA")), OrderBA := "Secondary"]
tabBileAcidUntgt[grep("(keto|iso)", Molecule, ignore.case = TRUE), OrderBA := "Secondary"]

# Upper-case-ify
tabBileAcidUntgt$Molecule <-
  tabBileAcidUntgt$Molecule %>% 
  Hmisc::capitalize()

# Define sort order based on WBF-011
orderUntgtMetabs <- 
  tabBileAcidUntgt %>% 
  .[(treatment == "wbf11")] %>% 
  setorder(-estimateWinGrp) %>% 
  .$Molecule
tabBileAcidUntgt[, metaboFactor := factor(Molecule, levels = orderUntgtMetabs)]
```

Define base plot

```{r fig.width=14}
pBileAcidUntgt <-
  tabBileAcidUntgt %>% copy %>% 
  .[(ConjugateFamily %in% showConjuFams)] %>% 
  .[, conjuFam := factor(ConjugateFamily, levels = showConjuFams)] %>% 
  # .[(pBwGrp < 0.05)] %>%
  ggplot(aes(metaboFactor, estimateWinGrp, color = treatment, fill = treatment)) +
  geom_hline(yintercept = 0.0, size = 0.1) +
  facet_wrap(
    facets = ~conjuFam, nrow = 1,
    scales = "free_x", shrink = TRUE, drop = TRUE,
    strip.position = "top") +
  # Adding this has the effect of centering each panel
  geom_blank(mapping = aes(ymin = -confIntLo, ymax = -confIntHi)) +
  # The (within-group) log2-ratio + C.I.s
  geom_pointrange(
    alpha = 1.0,
    size = 0.5,
    fatten = 3,
    shape = 23,
    stroke = 0.1,
    position = position_dodge(width = 0.6),
    mapping = aes(ymin = confIntLo, ymax = confIntHi)
  ) +
  ylab(
    expression(paste(
      log[2](over(Endpoint,Baseline))
    ))) +
  # xlab("")
  scaleColorManualTreatment +
  scaleFillManualTreatment +
  ggtitle("Plasma, untargeted") +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    axis.ticks = element_blank(),
    axis.ticks.x = element_line(size = 0.25),
    # strip.text = element_blank(),
    strip.background = element_blank(),
    axis.title.x = element_blank(),
    axis.text.x = element_text(
      size = 10,
      angle = 30, 
      hjust = 1, 
      vjust = 1))
pBileAcidUntgt

## Show UDCA precursors statistics in WBF-011
tabBileAcidUntgt %>% 
  .[grep("^(Glyco)*[Ll]ithocholic acid$", Molecule)] %>% 
  rbind(., 
        (tabBileAcidUntgt %>% 
           .[grep("^(Glyco)*[Cc]henodeoxycholic acid$", Molecule)])
  ) %>% 
  .[, .(Molecule, treatment, pBwGrp, estimateWinGrp, pWinGrp)] %>% 
  setorder(-pWinGrp) %>% 
  knitr::kable(digits = 2)

## Simplified, untargeted plasma bile acid changes
## Trim down to the less-exotic 
## (and usually higher concentration) molecules
filterStringBileAcids <- "(tauro|keto|iso|sulfate|glucuronide)"
pBileAcidUntgt$data <-
  pBileAcidUntgt$data[grep(filterStringBileAcids, Molecule, 
                           ignore.case = TRUE, invert = TRUE)]
pBileAcidUntgt

## Table
## Summarize key differences in a table.
pBileAcidUntgt$data[(pBwGrp < 0.05)][(pWinGrp < 0.1)][(treatment == "wbf11")]
# Define withing group marking symbol
pBileAcidUntgt$data[, symbolWinGrp := ""]
pBileAcidUntgt$data[(pBwGrp < 0.05 & pWinGrp < 0.1), symbolWinGrp := "."]
pBileAcidUntgt$data[(pBwGrp < 0.05 & pWinGrp < 0.05), symbolWinGrp := "*"]
# Define between-group (bracket) symbol
pBileAcidUntgt$data[, symbolBwGrp := ""]
pBileAcidUntgt$data[(pBwGrp < 0.05 & pWinGrp < 0.1 & treatment == "wbf11"), symbolBwGrp := "*"]
pBileAcidUntgt$data %>% 
  .[(pBwGrp < 0.05 & pWinGrp < 0.1)] %>% 
  .[, .(Molecule, ConjugateFamily, estimateWinGrp, pWinGrp, pBwGrp, symbolWinGrp)] %>% 
  knitr::kable(digits = 3)

## Add brackets
tabBileAcidUntgtBrkt <- NULL
tabBileAcidUntgtBrkt <- 
  pBileAcidUntgt$data %>% copy %>% 
  .[(pBwGrp < 0.05 & pWinGrp < 0.1)] %>% 
  .[, xpos := .I]
pBileAcidUntgt <-
  pBileAcidUntgt +
  # Bw-grp bracket
  ggpubr::geom_bracket(
    mapping = aes(
      y.position = confIntHi, 
      label = symbolBwGrp, 
      xmin = xpos - 0.2,
      xmax = xpos + 0.2),
    tip.length = c(0.2, 0.02),
    bracket.nudge.y = 0.4,
    color = "black",
    vjust = 0.3, 
    hjust = 0.5,
    data = tabBileAcidUntgtBrkt,
    label.size = 4,
  ) +
  # Within-group nominal stat sig
  geom_text(
    data = tabBileAcidUntgtBrkt,
    nudge_y = 0.1,
    mapping = aes(label = symbolWinGrp, y = confIntHi, x = xpos + 0.2)
  )
pBileAcidUntgt
```



# Fig. 3b Targeted Bile Acids

## Event-Wide Table

Compute subject-wise Week12, Baseline paired metrics

```{r}
loopAndIndicatorVars <- 
  c("treatment", "quantCategory", "detectCategory",
    "OrderBA", "ConjugateFamily", "Molecule")
formulaPivotEventWide <-
  loopAndIndicatorVars %>% 
  c("Subject", .) %>% 
  paste(collapse = " + ") %>% 
  paste0(" ~ Event", collapse = "") %>% 
  as.formula()

tabPlasmaTgtBileAcidsEventWide <- NULL
tabPlasmaTgtBileAcidsEventWide <-
  tabPlasmaMsOmicsBileAcidsLong %>%
  # pivot wide to ensure consistent handling of subject-event-sample and missing
  dcast.data.table(
    formula = formulaPivotEventWide,
    # formula = Treatment + Subject + 
    #   detectCategory + quantCategory + 
    #   ConjugateFamily + Molecule ~ Event, 
    fill = NA, 
    fun.aggregate = mean,
    value.var = "Concentration")
# How many different quantitative-accuracy bile acids were measured?
tabPlasmaTgtBileAcidsEventWide %>% 
  .[(quantCategory == "absolut")] %>% 
  .$Molecule %>% 
  unique() %>% 
  sort()

# Define concise table of Dummy values, set to LoD
tabLoDMsOmics <-
  tabPlasmaMsOmicsBileAcidsLong %>% 
  .[, .(Molecule, LOD)] %>% 
  unique()
# Add LoD back to each entry in 'wide' table
tabPlasmaTgtBileAcidsEventWide <- 
  tabLoDMsOmics[tabPlasmaTgtBileAcidsEventWide, on = "Molecule"]
tabPlasmaTgtBileAcidsEventWide[(Baseline < LOD), Baseline := LOD]
tabPlasmaTgtBileAcidsEventWide[(Week12 < LOD), Week12 := LOD]
# Compute change statistics
tabPlasmaTgtBileAcidsEventWide <-
  tabPlasmaTgtBileAcidsEventWide %>% 
  .[, Week12MinusBaseline := Week12 - Baseline]

## Test, "Delta": Week12 - Baseline
tabPlasmaTgtBileAcidsEventWide %>% 
  .[is.infinite(Week12MinusBaseline)] %>% 
  nrow()
tabPlasmaTgtBaWinGrp <- NULL
tabPlasmaTgtBaWinGrp <-
  tabPlasmaTgtBileAcidsEventWide %>% 
  .[, .(
    wilcoxOut = list(
      wilcox.test(
        x = Week12MinusBaseline, 
        alternative = "two.sided",
        conf.int = TRUE,
        mu = 0.0, 
        paired = FALSE)
    )
  ), 
  by = loopAndIndicatorVars]
# Extract key results from test
extract_wi = function(wilcoxOut){data.table(
  pvalue = wilcoxOut$p.value,
  statistic = wilcoxOut$statistic,
  estimate = wilcoxOut$estimate,
  confIntLo = wilcoxOut$conf.int[1],
  confIntHi = wilcoxOut$conf.int[2]
)}
tabPlasmaTgtBaWinGrp <-
  tabPlasmaTgtBaWinGrp[, extract_wi(wilcoxOut[[1]]), 
                                   by = loopAndIndicatorVars]
setorder(tabPlasmaTgtBaWinGrp, -pvalue, treatment)

## Between-Group: WBF-011 v. Placebo
# Compute wilcoxon
tabPlasmaTgtBaBwGrp <- NULL
# Compute the between-group comparison WBF-011 v. Placebo
tabPlasmaTgtBaBwGrp <-
  tabPlasmaTgtBileAcidsEventWide %>% 
  # .[(numNonMissing > 0)] %>% 
  # .[(numNonzeroLogRatio > thresholdMinNonzeroLogRatios)] %>%
  .[(treatment %in% c("wbf11", "placebo"))] %>% 
  .[, .(
    wilcoxOut = list(
      try(expr = {
        wilcox.test(
          x = Week12MinusBaseline[(treatment == "wbf11")],
          y = Week12MinusBaseline[(treatment == "placebo")],
          alternative = "greater",
          conf.int = TRUE,
          paired = FALSE)
      }, silent = TRUE)
    )
  ), 
  by = c(loopAndIndicatorVars[-1])]
# Extract key results from test
extract_wi = function(wilcoxOut){
  data.table(
    pvalue = wilcoxOut$p.value,
    statistic = wilcoxOut$statistic,
    estimate = wilcoxOut$estimate,
    confIntLo = wilcoxOut$conf.int[1],
    confIntHi = wilcoxOut$conf.int[2]
  )
}
# Check that test didn't fail and return non-sense (missing or funky data)
tabPlasmaTgtBaBwGrp[, success := inherits(wilcoxOut[[1]], "htest"), 
                           by = c(loopAndIndicatorVars[-1])]
tabPlasmaTgtBaBwGrp <-
  tabPlasmaTgtBaBwGrp %>% 
  .[(success), try({extract_wi(wilcoxOut[[1]])}, silent = TRUE), 
    by = c(loopAndIndicatorVars[-1])]
setorder(tabPlasmaTgtBaBwGrp, -pvalue)
tabPlasmaTgtBaBwGrp %>% .[(quantCategory == "absolut")] %>% .[(pvalue < 0.25)] %>% tail(10)
# Join the two comparison tables (within- and between-group)
setnames(tabPlasmaTgtBaWinGrp, "pvalue", "pWinGrp")
setnames(tabPlasmaTgtBaWinGrp, "estimate", "estimateWinGrp")
setnames(tabPlasmaTgtBaBwGrp, "pvalue", "pBwGrp")
setnames(tabPlasmaTgtBaBwGrp, "estimate", "estimateBwGrp")
tabPlasmaTgtWilcox <-
  (tabPlasmaTgtBaBwGrp %>% 
     .[, .(Molecule, estimateBwGrp, pBwGrp)]) %>% 
  .[tabPlasmaTgtBaWinGrp, on = "Molecule"] %>% 
  copy()
```

Show UDCA precursors statistics in WBF-011

```{r}
tabPlasmaTgtWilcox %>% 
  .[grep("^(Glyco)*[Ll]ithocholic acid$", Molecule)] %>% 
  .[(quantCategory == "absolut")] %>% 
  rbind(., 
        (tabPlasmaTgtWilcox %>% 
           .[grep("^(Glyco)*[Cc]henodeoxycholic acid$", Molecule)])
  ) %>% 
  .[, .(Molecule, treatment, pBwGrp, estimateWinGrp, pWinGrp)] %>% 
  setorder(-pWinGrp) %>% 
  knitr::kable(digits = 2)
```

Define plotting order, organization.

```{r}
# Define molecule order based on significance results 
showOrder <- showSet <- NULL
showSet <- 
  tabPlasmaTgtBaBwGrp %>% copy %>% 
  .[(quantCategory == "absolut")] %>% 
  # Select the groups to show in main plot
  .[(ConjugateFamily %in% showConjuFams)] %>% 
  # .[(pvalue < 0.5)] %>%
  .$Molecule
# Define order of labels (a grouping)
showOrder <- 
  tabPlasmaTgtWilcox %>% copy %>% 
  .[showSet, on = "Molecule"] %>% 
  .[(treatment == "wbf11")] %>% 
  setorder(-estimateWinGrp) %>% 
  .$Molecule
```

Targeted Summary, Delta

```{r}
# Define table for plot
tabPlasmaTargetedBasePlot <-
  tabPlasmaTgtWilcox %>% copy %>% 
  .[showSet, on = "Molecule"] %>% 
  # Make order match the untargeted results, via orderUntgtMetabs
  .[, bileAcids := factor(Molecule, 
                          levels = c(orderUntgtMetabs, "Lithocholic acid"))] %>%
  .[, conjuFam := factor(ConjugateFamily, levels = showConjuFams)]

# Define plot
pPlasmaTargetedBase <-
  tabPlasmaTargetedBasePlot %>% 
  ggplot(aes(x = bileAcids, y = estimateWinGrp)) +
  # facet_grid(ConjugateFamily ~ ., scales = "free") +
  facet_wrap(~conjuFam, scales = "free", nrow = 1) +
  geom_hline(yintercept = 0.0, size = 0.1) +
  geom_blank(mapping = aes(ymin = -confIntLo, ymax = -confIntHi)) +
  geom_pointrange(
    position = position_dodge(width = 0.6),
    alpha = 1.0,
    size = 0.5,
    fatten = 3,
    shape = 23,
    stroke = 0.1,
    mapping = aes(
      fill = treatment,
      color = treatment,
      ymin = confIntLo, 
      ymax = confIntHi
    )
  ) +
  scaleColorManualTreatment +
  scaleFillManualTreatment +
  # coord_flip() +
  ylab("\u0394 Concentration [\u03BCM]") +
  ggtitle("Plasma, targeted") +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    axis.ticks.x = element_line(size = 0.25),
    # panel.border = element_blank(),
    strip.background = element_blank(),
    axis.title.y = element_text(size = 14, angle = 90, vjust = 0.5),
    axis.title.x = element_blank(),
    axis.text.x = element_text(
      size = 10,
      angle = 30, 
      hjust = 1, 
      vjust = 1))
pPlasmaTargetedBase

## Simplified, targeted plasma bile acid changes
## Trim down to the less-exotic 
## (and usually higher concentration) molecules
pPlasmaTargetedBase$data <-
  pPlasmaTargetedBase$data[grep(filterStringBileAcids, Molecule, 
                                ignore.case = TRUE, invert = TRUE)]
pPlasmaTargetedBase

## Table
## Summarize key differences in a table.
pPlasmaTargetedBase$data[(pBwGrp < 0.15)][(pWinGrp < 0.3)][(estimateWinGrp > 0.0)]
# Define withing group marking symbol
pPlasmaTargetedBase$data[, symbolWinGrp := ""]
pPlasmaTargetedBase$data[(pBwGrp < 0.15 & pWinGrp < 0.3 & estimateWinGrp > 0.0), symbolWinGrp := "."]
# pPlasmaTargetedBase$data[(pBwGrp < 0.05 & pWinGrp < 0.05), symbolWinGrp := "*"]
# Define between-group (bracket) symbol
pPlasmaTargetedBase$data[, symbolBwGrp := ""]
pPlasmaTargetedBase$data[(pBwGrp < 0.15 & pWinGrp < 0.3 & 
                            estimateWinGrp > 0.0 & 
                            treatment == "wbf11"), symbolBwGrp := "."]
pPlasmaTargetedBase$data %>% 
  .[(pBwGrp < 0.15 & pWinGrp < 0.3 & 
       estimateWinGrp > 0.0)] %>% 
  .[, .(Molecule, ConjugateFamily, estimateWinGrp, pWinGrp, pBwGrp, symbolWinGrp)] %>% 
  knitr::kable(digits = 3)

## Add brackets
tabBileAcidTgtBrkt <- NULL
tabBileAcidTgtBrkt <- 
  pPlasmaTargetedBase$data %>% copy %>% 
  .[(symbolBwGrp != "")] %>% 
  .[, xpos := .I]
pPlasmaTargeted <-
  pPlasmaTargetedBase +
  # Bw-grp bracket
  ggpubr::geom_bracket(
    mapping = aes(
      y.position = confIntHi, 
      label = symbolBwGrp, 
      xmin = xpos - 0.2,
      xmax = xpos + 0.2),
    tip.length = c(0.2, 0.1),
    bracket.nudge.y = 0.04,
    color = "black",
    vjust = 0.15, 
    hjust = 0.5,
    data = tabBileAcidTgtBrkt,
    label.size = 6,
  ) +
  # Within-group nominal stat sig
  # (nothing: within-group sig weaker than the untargeted log-ratio data)
  geom_text(
    data = tabBileAcidTgtBrkt[(pWinGrp < 0.1)],
    nudge_y = 0.01,
    mapping = aes(
      color = treatment,
      label = symbolWinGrp, 
      y = confIntHi, 
      x = xpos + 0.2)
  )
pPlasmaTargeted
```


# Fig. 3c Total bile acids

For completeness, sum up total bile acids,
show whether it is within healthy range for all subjects.

- Annotated the reference ranges.
- Annotate the medians within group, and grand at each timepoint.

```{r}
studyArmDodgeWidth = 0.8
thresholdMaxTotalPlot = 15

tabShowBileAcids <- pPlasmaTotalBileAcids <- NULL
tabShowBileAcids <- 
  tabPlasmaMsOmicsBileAcidsLong %>% copy %>% 
  .[(
    detectCategory == "Final results" &
      quantCategory == "absolut")] %>% 
  .[, .(totalBileAcids = sum(Concentration, na.rm = TRUE)), 
    by = .(treatment, Subject, Event)] %>% 
  .[, Treatment := studyArmPretty[treatment]]
tabPlotTotalBileAcids <-
  tabShowBileAcids %>% copy %>% 
  .[(totalBileAcids > thresholdMaxTotalPlot),
    totalBileAcids := Inf]
pPlasmaTotalBileAcids <- 
  tabPlotTotalBileAcids %>% 
  ggplot(aes(Treatment, totalBileAcids, color = Treatment, fill = Treatment)) +
  facet_wrap(~Event, strip.position = "top") +
  # Normal range, 0 - 10 uM
  # Moderate cholestasis ~10 - 40 uM
  geom_rect(
    data = data.frame(Treatment = "dummy", Subject = "SS"),
    color = NA,
    fill = "darkred",
    alpha = 0.15,
    inherit.aes = FALSE,
    xmin = -Inf, 
    xmax = Inf,
    ymin = 10,
    ymax = Inf
  ) +
  # severe cholestasis
  # >= 40 uM
  # (not shown)
  #
  # Baseline grand median
  geom_hline(
    yintercept = tabPlotTotalBileAcids[(Event == "Baseline")] %>% 
      .$totalBileAcids %>% 
      median(),
    size = 0.25,
    linetype = 1,
    alpha = 0.65,
  ) +
  geom_beeswarm(
    data = tabPlotTotalBileAcids[is.finite(totalBileAcids)],
    shape = 21,
    dodge.width = studyArmDodgeWidth,
    # mapping = aes(color = Treatment, fill = Treatment),
    groupOnX = TRUE,
    cex = 3.5,
    size = 1,
    stroke = 0.1,
    alpha = 0.75
  ) +
  # Annotate any points that get off the top of the screen
  ggrepel::geom_text_repel(
    position = position_dodge(width = studyArmDodgeWidth),
    data = tabShowBileAcids[(totalBileAcids > thresholdMaxTotalPlot)], 
    mapping = aes(
      y = Inf, 
      label = round(totalBileAcids, digits = 1)
    ),
    min.segment.length = 0,
    size = 2
  ) +
  # group-median crossbar
  stat_summary(
    geom = "crossbar",
    fatten = 1.3,
    width = 0.5,
    position = position_dodge(width = studyArmDodgeWidth),
    fun.data = function(x){
      thisMedian = median(x, na.rm = TRUE)
      return(
        data.frame(
          y = thisMedian,
          ymin = thisMedian,
          ymax = thisMedian
        )
      )
    }) +
  scaleColorTreatmentManPretty +
  scaleFillTreatmentManPretty +
  guides(
    color = guide_legend(title="Study Arm")
  ) +
  ylab("Total bile acids [\u03BCM]") +
  theme(
    panel.grid = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_line(size = 0.25),
    axis.text.x = element_text(angle = 30, size = 8, hjust = 1),
    legend.position = "none"
  )
pPlasmaTotalBileAcids
```


# Fig. 3e Monoculture $\Delta$ Bile Acids

## Load data

```{r}
tabStrainCultureMsOmicsBileAcids <-
  readRDS(params$tabStrainCultureMsOmicsBileAcids)
```

Check NIC

```{r}
tabStrainCultureMsOmicsBileAcids %>% 
  .[(inoculated == "FALSE")] %>% 
  .[(sampleType == "supernatant")] %>% 
  .[(quantCategory == "absolut")] %>% 
  ggplot(aes(timepoint, Concentration, color = ConjugateFamily)) +
  facet_grid(cols = vars(moleculeAdded), rows = vars(run)) +
  geom_hline(yintercept = 50, linetype = 3, size = 0.25) +
  geom_point() +
  theme(axis.ticks.y = element_line(size = 0.25)) +
  ggtitle("NIC only", "Looks good. No changes. Confirms design.")
```

Check growth-typical conditions (no added bile acids) controls.
(note: this control was not included in CBUT run)

```{r}
tabStrainCultureMsOmicsBileAcids %>% copy %>% 
  .[(Condition == "GrowthCtl")] %>% 
  .[(sampleType == "supernatant")] %>% 
  .[(quantCategory == "absolut")] %>% 
  ggplot(aes(timepoint, Concentration, color = ConjugateFamily)) +
  facet_grid(cols = vars(Strain)) +
  ylim(0, 50) +
  geom_path(mapping = aes(group = paste(Condition, Molecule))) +
  geom_point() +
  theme(
    axis.ticks.y = element_line(size = 0.25), 
    legend.position = "bottom") +
  ggtitle("Control: No primary bile acids amended", 
          "Minimal changes, low concentrations. Confirms design.")
```

Note: Should include only "absolut" concentration values in the figure-of-merit.

## Effective volume, umoles

Compute the effective-volume that combines
the cell pellet and supernatant measurements at `t_final`.

```{r}
tabMonocultureDeltaSummary <-
  tabStrainCultureMsOmicsBileAcids %>% copy %>%
  .[(inoculated == "TRUE")] %>%
  .[(Condition != "GrowthCtl")] %>% 
  .[(media == "pyg")] %>% 
  .[(quantCategory == "absolut")] %>% 
  # This sums together cell-pellet and supernatant at t_final,
  # on both micromoles and effective_volume values
  dcast.data.table(
    formula = inoculated + moleculeAdded + Strain +
      Condition +
      ConjugateFamily + OrderBA + Molecule ~ timepoint,
    value.var = c("micromoles", "effectiveVolume"),
    fun.aggregate = sum,
    na.rm = TRUE
  ) %>%
  # Compute effective concentrations at zero, final
  .[, conc0 := (1000 * micromoles_T0 / effectiveVolume_T0)] %>%
  .[, concf := (1000 * micromoles_Tfinal / effectiveVolume_Tfinal)] %>%
  # Deltas
  .[, deltaMicroMoles := micromoles_Tfinal -  micromoles_T0] %>%
  .[, deltaConc := concf - conc0]
```

Show each timepoint separately.

```{r}
tabPlotBothTimepoints <- 
  tabStrainCultureMsOmicsBileAcids %>% 
  .[!is.na(Strain)] %>% 
  .[(inoculated == "TRUE")] %>%
  .[(Condition != "GrowthCtl")] %>% 
  .[(media == "pyg")] %>% 
  .[(quantCategory == "absolut")] %>% 
  # Sum
  .[, .(micromoles = sum(micromoles)), 
    by = .(Strain, moleculeAdded, Condition, 
           ConjugateFamily, Molecule, timepoint)]
# Only show molecules that have micromoles > threshold
showMolecules <- tabPlotBothTimepoints[(micromoles > 1e-3)]$Molecule %>% unique()
tabPlotBothTimepoints %>% 
  .[showMolecules, on = "Molecule"] %>% 
  ggplot(aes(timepoint, micromoles, 
             # shape = sampleType,
             color = ConjugateFamily)) +
  facet_grid(
    cols = vars(moleculeAdded), 
    rows = vars(Strain)) +
  geom_path(mapping = aes(group = paste(Strain, Condition, Molecule))) +
  geom_point() +
  ggtitle("Inoculated only", "Replicates similar. Confirms design.")
```

Summarize as effective concentration deltas.

only `abs(\u0394) > 1\u03BCM"`

CBUT monoculture main bile acid changes
following growth in rich media amended with
CA, CDCA, or sterile water (control)

```{r}
pMonocultureDeltaSummary <- NULL
strainOrder <- c("CBUT", "EHAL", "CBEI", "AMUC", "BINF")
tabDeltaSummary <-
  tabMonocultureDeltaSummary %>% copy %>% 
  .[(Molecule %in% names(showBileAcids))] %>%
  # Pretty plotting
  .[, Amendment := paste0("+50 \u03BCM ", moleculeAdded)] %>% 
  # Omit the vanishing trace bile acids from this plot
  .[showMolecules, on = "Molecule", nomatch = 0] %>% 
  # set factor to order the stain facets
  .[, strainFac := factor(Strain, levels = strainOrder)] %>% 
  # limit negative-delta outlier effect on plot limits
  .[(deltaConc < -75), deltaConc := -Inf]
pMonocultureDeltaSummary <-
  tabDeltaSummary %>% 
  .[!is.na(Strain)] %>% 
  ggplot(
    mapping = aes(
      x = Molecule, 
      y = deltaConc, 
      color = OrderBA, 
      fill = OrderBA)
  ) +
  scale_y_continuous(breaks = c(-50, -25, 0, 25)) +
  facet_grid(
    rows = vars(Amendment), 
    cols = vars(strainFac)
  ) +
  geom_hline(yintercept = 0.0, size = 0.25, color = "black") +
  # The bars
  stat_summary(
    width = 0.5,
    alpha = 0.7,
    geom = "col",
    color = NA,
    fun.data = function(x){
      data.frame(y = median(x))
    }) +
  # The individual replicate values
  geom_point(
    size = 1.5,
    stroke = 0,
    alpha = 0.7,
    position = position_beeswarm(cex = 2.5, groupOnX = TRUE)
  ) +
  guides(
    colour = guide_legend("Bile Acid Rank"), 
    fill = guide_legend("Bile Acid Rank")) +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  # scale_y_continuous(breaks = c(-40, -20, 0, 20)) +
  ylab("\u0394 Concentration [\u03BCM]") +
  # ggtitle("\u0394 Concentration [\u03BCM]") +
  theme(
    plot.title = element_blank(),
    strip.text.x = element_text(size = 12),
    strip.text.y = element_text(size = 10, hjust = 0),
    legend.position = "none",
    panel.grid.major.y = element_line(size = 0.1, color = "gray"),
    axis.ticks = element_blank(),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 30, hjust = 1, size = 10),
    axis.title.x = element_blank(),
  ) 
# pMonocultureDeltaSummary
```

## Fig. 3e Monoculture $\Delta$ Summary

```{r}
pMonocultureDeltaSummary
```

For plot, make a version that only shows 
the targeted assay positive confirmation (CBUT).
Can leave as "data not shown" for now for the others,
since they are all UDCA-negligible.

```{r}
pMonocultureDeltaSummaryCbutOnly <- copy(pMonocultureDeltaSummary)
pMonocultureDeltaSummaryCbutOnly$data <-
  pMonocultureDeltaSummary$data[(Strain == "CBUT")] %>% 
  copy()
# Flip the facet grid for better vertical space usage
pMonocultureDeltaSummaryCbutOnly <- 
  pMonocultureDeltaSummaryCbutOnly +
  facet_grid(
    cols = vars(Amendment), 
    rows = vars(strainFac)
  )
pMonocultureDeltaSummaryCbutOnly
```



# Fig. 3d Strain Culture Pilot, Metabolon

Strain culture rich media + primary bile acids (CA + CDCA) pilot.
Metabolon untargeted data.

- 50 uM CA amended
- 50 uM CDCA amended
- No bile acid amendment for AMUC

```{r}
tabStrainCultureMetabolonPivWideType <- 
  # Uses Metabolon `OrigScale` rather than the rescaled, imputed values.
  # Has benefit of explicit missing values, and consistent peak-area relationship
  readRDS(params$tabStrainCultureMetabolonPivWideType)
tabStrainCultureMetabolon <- 
  readRDS(params$tabStrainCultureMetabolon)
# Subset to bile acids on the 'sampletype wide' table
vecBileAcidIds <- 
  tabStrainCultureMetabolon[grep("Bile", `SUB PATHWAY`)] %>% 
  .$COMP_ID %>% unique()
# Define bile acids table
tabStrainCultureMetabolonBileAcids <- NULL
tabStrainCultureMetabolonBileAcids <-
  tabStrainCultureMetabolonPivWideType %>% copy %>% 
  .[vecBileAcidIds, on = "COMP_ID"] %>% 
  # One is a very small value on this `OrigScale` scale
  .[is.na(CellPellet), CellPellet := 1.0] %>% 
  .[is.na(BlankMedia), BlankMedia := 1.0] %>% 
  .[is.na(Supernatant), Supernatant := 1.0] %>% 
  # Compute log-ratios
  .[, CpOverBlank := log10(CellPellet / BlankMedia)] %>% 
  .[, SuOverBlank := log10(Supernatant / BlankMedia)] %>% 
  # Melt (pivot long) so that can be combined
  melt.data.table(
    id.vars = c("COMP_ID", "Strain"), 
    measure.vars = c("CpOverBlank", "SuOverBlank"),
    variable.name = "SampleType", 
    value.name = "log10Ratio") %>% 
  .[, SampleType := gsub("SuOverBlank", "Supernatant", SampleType)] %>% 
  .[, SampleType := gsub("CpOverBlank", "CellPellet", SampleType)]

tabStrainCultureMetabolonBileAcids <-
  tabStrainCultureMetabolonBileAcids %>% 
  # Add back annotations
  (tabStrainCultureMetabolon %>%
     .[, .(COMP_ID, BIOCHEMICAL, `SUPER PATHWAY`, `SUB PATHWAY`,
           MASS, CAS, KEGG, `Group HMDB`)] %>%
     unique())[., on = "COMP_ID", nomatch = 0]
# Standardize nomenclature, and categorize conjugate family
tabStrainCultureMetabolonBileAcids[, Molecule := copy(BIOCHEMICAL)]
tabStrainCultureMetabolonBileAcids[, Molecule := gsub("late", "lic acid", Molecule)]
tabStrainCultureMetabolonBileAcids[, ConjugateFamily := "Other"]
# Define conjugate families
tabStrainCultureMetabolonBileAcids[grep("^([Tt]auro|[Gg]lyco)*(3-dehydro)*[Cc]holic acid", Molecule), 
                                   ConjugateFamily := "CA"]
tabStrainCultureMetabolonBileAcids[grep("^([Z7]-)*([Kk]eto|[Tt]auro|[Gg]lyco)*[Dd]eoxycholic acid", Molecule), 
                                   ConjugateFamily := "DCA"]
tabStrainCultureMetabolonBileAcids[grep("^([Tt]auro|[Gg]lyco)*(3-dehydro)*[Cc]hen(o)*deoxycholic acid", Molecule), 
                                   ConjugateFamily := "CDCA"]
tabStrainCultureMetabolonBileAcids[grep("^([Ii]so)*([Tt]auro|[Gg]lyco)*[Uu]rsodeoxycholic acid", Molecule), 
                                   ConjugateFamily := "UDCA"]
tabStrainCultureMetabolonBileAcids[grep("^([Tt]auro|[Gg]lyco)*[Uu]rsocholic acid", Molecule),
                                   ConjugateFamily := "UCA"]
tabStrainCultureMetabolonBileAcids[grep("[lL]ithocholic acid", Molecule), 
                                   ConjugateFamily := "LCA"]
tabStrainCultureMetabolonBileAcids[grep("[Hh]yocholic acid", Molecule), 
                                   ConjugateFamily := "HCA"]
tabStrainCultureMetabolonBileAcids[grep("[Hh]yodeoxycholic acid", Molecule), 
                                   ConjugateFamily := "HDCA"]
tabStrainCultureMetabolonBileAcids[grep("muri", Molecule, ignore.case = TRUE), 
                                   ConjugateFamily := "MCA"]
tabStrainCultureMetabolonBileAcids[, OrderBA := "Z"]
tabStrainCultureMetabolonBileAcids[(ConjugateFamily %in% c("CA", "CDCA", "HCA")), 
                                   OrderBA := "Primary"]
tabStrainCultureMetabolonBileAcids[(ConjugateFamily %in% 
                                      c("DCA", "LCA", "UDCA", "UCA", "HDCA")), 
                                   OrderBA := "Secondary"]
tabStrainCultureMetabolonBileAcids[grep("(keto|iso)", Molecule, ignore.case = TRUE), 
                                   OrderBA := "Secondary"]
# For interpretive organization, it helps to group these together, last
tabStrainCultureMetabolonBileAcids[grep("3-dehydro", Molecule),
                                   ConjugateFamily := "z3dehydro"]
tabStrainCultureMetabolonBileAcids[grep("3-dehydro", Molecule),
                                   OrderBA := "Secondary"]

# Upper-case-ify
tabStrainCultureMetabolonBileAcids$Molecule <-
  tabStrainCultureMetabolonBileAcids$Molecule %>% 
  Hmisc::capitalize()

# Define order of metabolites
orderStrainCultureMetabolonBileAcids <-
  tabStrainCultureMetabolonBileAcids %>% 
  .[, .(OrderBA, ConjugateFamily, Molecule)] %>% 
  unique() %>% 
  setorder(-OrderBA, -ConjugateFamily, -Molecule) %>% 
  # show()
  .$Molecule

# Define strain culture pilot summary heatmap
pStrainCultureMetabolonBileAcids <-
  tabStrainCultureMetabolonBileAcids %>% copy %>% 
  .[, MoleculeFac := factor(Molecule, 
                            levels = orderStrainCultureMetabolonBileAcids)] %>% 
  .[, SampleTypePretty := c(CellPellet = "Cell Pellet", 
                            Supernatant = "Supernatant")[SampleType]] %>% 
  ggplot(aes(SampleTypePretty, MoleculeFac, fill = log10Ratio)) +
  # facet_grid(ConjugateFamily ~ Strain, shrink = FALSE, drop = TRUE, scales = "free") +
  facet_grid( ~ Strain, shrink = FALSE, drop = TRUE, scales = "free") +
  geom_raster() +
  scale_fill_gradient2() +
  # Adjust the color mapping legend
  guides(
    fill = guide_colourbar(
      title = expression(paste(
        log[10](over(Specimen,Medium))
      )),
      # nbin = 5, 
      barwidth = 13, 
      barheight = 0.5)
  ) +
  theme(
    legend.position = "bottom",
    legend.justification = "left",
    legend.title = element_text(size = 8, vjust = 0.2),
    axis.title = element_blank(),
    axis.text.x = element_text(angle = 30, size = 10, hjust = 1)
  )
pStrainCultureMetabolonBileAcids

pStrainCultureMetabolonBileAcids$data <- 
  # Drop AMUC, because no primary bile acids amended
  pStrainCultureMetabolonBileAcids$data %>% 
  .[(Strain != "AMUC")]
pStrainCultureMetabolonBileAcids
```


# Build Figure 3

```{r pFig3, fig.height=7, fig.width=6.7}
fig2Layout <- "
AAAAAAAA#
BBBBBBBCC
DDDD##EEE
"
sizeAxisTitles <- 7
sizeAxisText <- 6
pFig3 <- NULL
pFig3 <-
  # Untargeted, Log-Ratio
  (
    pBileAcidUntgt + 
      theme(
        panel.spacing = unit(8, "pt"),
        axis.ticks.y = element_line(size = 0.25),
        plot.title =  element_text(size = sizeAxisTitles),
        strip.text = element_text(size = sizeAxisTitles),
        axis.title.x = element_blank(),
        axis.text.x = element_text(size = sizeAxisText),
        axis.text.y = element_text(size = sizeAxisText),
        axis.title.y = element_text(
          size = sizeAxisTitles, angle = 0, vjust = 0.5,
          margin = margin(r = -100, unit = "pt"))
      )
  ) +
  # Targeted, Delta
  (
    pPlasmaTargeted + 
      theme(
        panel.spacing = unit(1, "pt"),
        plot.title =  element_text(size = sizeAxisTitles),
        strip.text = element_text(size = sizeAxisTitles),
        axis.title.x = element_blank(),
        axis.text.x = element_text(size = sizeAxisText),
        axis.text.y = element_text(size = sizeAxisText),
        axis.title.y = element_text(
          size = sizeAxisTitles, angle = 90, vjust = 0.5,
          margin = margin(r = -100, unit = "pt"))
      )
  ) +
  # Targeted, total bile acids
  (pPlasmaTotalBileAcids + 
     theme(
       panel.spacing = unit(1, "pt"),
       plot.margin = margin(0, unit='pt'),
       plot.title =  element_text(size = sizeAxisTitles),
       strip.text = element_text(size = sizeAxisTitles),
       axis.title.x = element_blank(),
       axis.title.y = element_text(size = sizeAxisTitles),
       axis.text.x = element_text(size = sizeAxisText),
       axis.text.y = element_text(size = sizeAxisText)
     )
  ) +
  # In vitro monoculture pilot summary heatmap
  (
    pStrainCultureMetabolonBileAcids +
      # Adjust the color mapping legend
      guides(
        fill = guide_colourbar(
          title = expression(paste(
            log[10](over(Specimen,Medium))
          )),
          barwidth = 10, 
          barheight = 0.5)
      ) +
      theme(
        panel.spacing = unit(1, "pt"),
        plot.title =  element_text(size = sizeAxisTitles),
        strip.text = element_text(size = sizeAxisTitles),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_text(size = sizeAxisText),
        axis.text.y = element_text(size = sizeAxisText),
        legend.text = element_text(size = sizeAxisText), 
        legend.title = element_text(size = sizeAxisText, vjust = 0.2),
        legend.margin = margin(l = -80, r = 200, t = -25, unit='pt')
      )
  ) +
  (pMonocultureDeltaSummaryCbutOnly + 
     theme(
       plot.margin = margin(0, unit='pt'),
       axis.ticks.y = element_line(size = 0.25),
       axis.ticks.x = element_line(size = 0.25),
       plot.title = element_blank(),
       strip.text.x = element_text(size = sizeAxisTitles),
       strip.text.y = element_text(size = sizeAxisTitles),
       axis.title.x = element_blank(),
       axis.title.y = element_text(size = sizeAxisTitles),
       axis.text.x = element_text(size = sizeAxisText),
       axis.text.y = element_text(size = sizeAxisText)
     )
  ) +
  plot_annotation(tag_levels = 'a')
# render prototype in report
pFig3 +
  plot_layout(design = fig2Layout) & 
  theme(plot.tag = element_text(size = 8, face = "bold"))
```

```{r ggsaves}
figHeight = 180
ggsave(
  "Figure-03.png", 
  pFig3 +
    plot_layout(design = fig2Layout) & 
    theme(plot.tag = element_text(size = 8, face = "bold")), 
  width = 10.5, height = 10)
ggsave("Figure-03.pdf", 
       pFig3 +
         plot_layout(design = fig2Layout) & 
         theme(plot.tag = element_text(size = 8, face = "bold", family = "Sans")), 
       device = cairo_pdf,
       width = 170,
       dpi = 300, 
       height = figHeight, 
       units = "mm")
```

